---
title: "Algorithm Demos"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Algorithm Demos}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
if(requireNamespace("pkgload", quietly = TRUE)) {
  pkgload::load_all(".", quiet = TRUE)
} else if(requireNamespace("immutables", quietly = TRUE)) {
  library(immutables)
} else {
  stop("Need either installed 'immutables' or the 'pkgload' package to render this vignette.")
}
source(file.path("helpers", "algorithm_demos_helpers.R"), local = TRUE)
```

## About these demos

The demos below illustrate the types of algorithms supported by `immutables`,
especially `interval_index()` as applied to problems in computational geometry.
They are not intended as how-to guides; the underlying scripts are complex,
largely AI generated, and 
mostly composed of plotting and animation machinery.

Each section gives a picture of the algorithm, names the structures it uses,
and shows a pre-rendered animation of a run. The animations are not produced
during vignette build; they are GIFs under
`vignettes/assets/algorithm-demos/`. Source paths for every demo and the asset
renderer are collected in a single section at the end.

<details>
<summary>Maintainer note: re-render one demo asset</summary>

```r
source("vignettes/helpers/render_algorithm_demo_assets.R")
render_algorithm_demo_assets(demos = "sweep_line")
```

Or from the shell:

```sh
R -q -f vignettes/helpers/render_algorithm_demo_assets.R --args demo=sweep_line
```

</details>

## Sweep line

Many algorithms in computational geometry utilize a sweep line. In this strategy
points belonging to geometric objects (e.g. line segments) are ordered along
one dimension, and then iterated over. Each point triggers some action, for 
example adding or removing the associated object from an "active" set and 
performing some computation on that active set.

In this very basic example, a sweep line loops over all line segments in an 
`interval_index()`. It collects all segments into an `index`, and loops over 
distinct endpoint values, using the `match_at` parameter of `pop_all_point()` to
identify segments to move from the index to an active set (those that start at the
current sweep line position) and remove from the active set (those end at 
the current sweepline position). 

It also keeps a record by appending the active set to a `playback` flexseq. 
Because these structures are immutable, we needn't worry about the active set
changing as the algorithm proceeds.

```{r}
items  <- c("a", "b", "c", "d")
starts <- c( 1,   2,   3,   4)
ends   <- c( 3,   5,   5,   6)

# we could also use xs <- sort(unique(c(starts, ends)))
xs <- as_ordered_sequence(c(starts, ends), keys = c(starts, ends))

index <- as_interval_index(items, start = starts, end = ends)

active <- interval_index()
playback <- flexseq()

# loop(for(...) {...}) is needed to loop over an immutables structure
loop(for(x in xs) {
  arrived <- pop_all_point(index, x, match_at = "start")
  index <- arrived$remaining
  active <- merge(active, arrived$elements)

  retired <- pop_all_point(active, x, match_at = "end")
  active <- retired$remaining

  cat("At x =", x, "active:", paste(unlist(active), collapse = ", "), "\n")

  playback <- push_back(playback, active)
})
```
Here's an illustration with more ranges, higlighting the sweep line position
and active set:

```{r sweep-line-viz, echo=FALSE, fig.cap="Sweep-line interval activity over persistent snapshots."}
knitr::include_graphics(algorithm_demo_asset("sweep_line"))
```

## Segment sweep

In this variant we discover pairwise intersections among line segments in the
2-D plane.

The event stream (segment start/end events) is an
`ordered_sequence` keyed by x. Active segment candidates are queried from an
`interval_index` built on segment x-ranges. A *second*
`ordered_sequence` accumulates discovered intersections as the sweep
progresses. A `flexseq` holds the snapshot
timeline, so the animation can replay the sweep with the intersection list
growing step by step.

```{r segment-sweep-viz, echo=FALSE, fig.cap="Segment sweep with active-set intersection discovery."}
knitr::include_graphics(algorithm_demo_asset("segment_sweep"))
```

## Convex hull

Computing convex hull also utilizes a sweep line method, or rather a pair of sweeps,
at least for [Andrew's monotone chain](https://en.wikibooks.org/wiki/Algorithm_Implementation/Geometry/Convex_hull/Monotone_chain) method.
After sorting the points by $x$ value, we sweep once to build the
lower hull, and then sweep again for the upper hull. Each sweep uses a stack: 
push the next point, then pop while the last three points
turn the wrong way.

Here `flexseq` carries most of the work. The sorted
input lives in an `ordered_sequence`, but the lower and upper hull stacks are
themselves `flexseq`s, and every push and pop returns a new one rather than
mutating in place. A third `flexseq` stores the snapshot timeline, so after
the run we have a complete record of every intermediate hull for visualization, 
including which point was pushed, which points were popped because they violated
convexity, and what the partial hull looked like at every step.

```{r convex-hull-viz, echo=FALSE, fig.cap="Monotone-chain hull construction with persistent snapshots."}
knitr::include_graphics(algorithm_demo_asset("convex_hull"))
```

## Fortune frontier

Fortune's algorithm](https://en.wikipedia.org/wiki/Fortune%27s_algorithm) is
a sweep-line method for efficiently computing Voronoi diagrams. A vertical
sweep moves across the set of points; each new site introduces a parabola to
the "beach line" frontier, and the breakpoints between adjacent parabolas
trace out the Voronoi edges as the sweep advances.

Three structures share the work. A `priority_queue` holds events keyed by x
and `pop_min()` produces the next event due. Two event types flow through it:
*site events* (one per input point, known up front) and *circle events*
(emerging dynamically whenever three adjacent arcs on the beach line begin
converging on a circumcenter; processing one creates a Voronoi vertex and
retires the middle arc). The active arcs along the frontier live in an
`ordered_sequence`, keyed by a composite score so they stay in frontier
order. A `flexseq` accumulates the *breakpoint trail*--every Voronoi edge
segment traced out so far--and a second `flexseq` holds the snapshot timeline.
The trail is an especially good fit for persistence: it only ever grows, and
the accumulated value from any past step is still available for comparison
or partial rendering with no bookkeeping. This animation samples intermediate
sweep positions for smoothness; structurally, the sweep only needs to stop at
site and circle events.

```{r fortune-viz, echo=FALSE, fig.cap="Fortune-style frontier sweep and revealed Voronoi structure."}
knitr::include_graphics(algorithm_demo_asset("fortune_frontier"))
```

## A* search

[A* search](https://en.wikipedia.org/wiki/A*_search_algorithm) finds a
shortest path from a start cell to a goal cell on a weighted graph by
always expanding the frontier node with the smallest estimated total cost.
Each node `n` carries two numbers: `g(n)`, the known cost to reach it from
the start, and `h(n)`, a heuristic estimate of the remaining cost to the
goal. The algorithm repeatedly pops the node with the smallest `f = g + h`
from a frontier, marks it visited, and updates its neighbors with any 
newly-cheaper path estimates. With an admissible heuristic this is guaranteed 
to find an optimal path while typically exploring far fewer nodes than 
Dijkstra's algorithm. 

This demo runs A* on a generated maze from a fixed start to a goal.
Three structures share the work. A `priority_queue` keyed by f-score holds
the frontier, and `pop_min()` picks the next node to expand. A separate
`ordered_sequence` keyed by node index records which nodes have already
been expanded, so the "already visited?" check is a `peek_key()`. 
A `flexseq` accumulates a snapshot of the frontier, the set of
expanded nodes, and the current node at every step. Since the full search
history is preserved as an immutable timeline, a post-hoc aggregate like
*pressure*, i.e. how many times each cell appeared on the frontier over
the course of the search, can be computed by folding over the snapshots
after the run. The animation exploits this to reveal the final path on
top of a pressure heatmap once the search settles.

```{r astar-viz, echo=FALSE, fig.cap="A* replay with staged final-route reveal, then persistent pressure summary."}
knitr::include_graphics(algorithm_demo_asset("astar_organic_history_large_fast"))
```

## A* with heuristic trail

A variation on the A* search above, run over a procedurally-generated
weighted terrain grid and augmented with a lagged retirement "clear wave"
that fades expanded cells out of the visualization instead of leaving them
permanently marked. The search itself is standard weighted A*; the clear
wave is visualization machinery with structural significance.

Five roles share the work. The A* frontier is still a `priority_queue`
keyed by `f = g + h`, and the expanded set is still an `ordered_sequence`
keyed by cell index. Running alongside them: a second `priority_queue`
holds scheduled retirement events, and a second `ordered_sequence` tracks
the currently-visible trail. A `flexseq` accumulates snapshots of all of
these at every expansion step.

What makes the trail interesting is how the retirement lag is computed.
Each expanded cell `c` is scheduled to clear at
`step_expand + lag_base + lag_warp * h(c)`: cells far from the goal (high
`h`) linger in the trail longer, while cells close to the goal clear
quickly. The tail is therefore thick in high-heuristic regions the
search recently moved through, and thin near the goal where the search
is narrowing in. Each iteration of the loop pops matured retirements
from the clear queue before snapshotting, so the frame drawn at every
step reflects exactly what should currently be visible.

```{r astar-trail-viz, echo=FALSE, fig.cap="A* expansion with a heuristic-warped clear wave trailing behind."}
knitr::include_graphics(algorithm_demo_asset("astar_trail"))
```

## Source and rendering

The full implementations live under `vignettes/helpers/demos/`:

- `demo_sweep_line_immutables.R`
- `demo_segment_sweep_immutables.R`
- `demo_convex_hull_immutables.R`
- `demo_fortune_frontier_immutables.R`
- `demo_astar_immutables.R`
- `demo_astar_trail_immutables.R`

The vignette-side helpers (asset paths, render presets, the A* preset used by
the large-maze animation) are in
`vignettes/helpers/algorithm_demos_helpers.R`, and
`vignettes/helpers/render_algorithm_demo_assets.R` is the entry point for
re-rendering any or all of the GIFs.
