---
title: "Algorithm Demos"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Algorithm Demos}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
if(requireNamespace("pkgload", quietly = TRUE)) {
  pkgload::load_all(".", quiet = TRUE)
} else if(requireNamespace("immutables", quietly = TRUE)) {
  library(immutables)
} else {
  stop("Need either installed 'immutables' or the 'pkgload' package to render this vignette.")
}
source(file.path("helpers", "algorithm_demos_helpers.R"), local = TRUE)
```

## About these demos

The examples below are not step-by-step
tutorials: the underlying scripts are several hundred to a thousand lines each,
but most of that code is event generation, plotting, and animation machinery.
What matters here is the *shape* of each algorithm — which structure carries
which role, and what persistence specifically buys when that role is filled by
an immutable value instead of a mutable one.

Each section gives a picture of the algorithm, names the structures it uses,
and shows a pre-rendered animation of a run. The animations are not produced
during vignette build; they are GIFs under
`vignettes/assets/algorithm-demos/`. Source paths for every demo and the asset
renderer are collected in a single section at the end.

<details>
<summary>Maintainer note: re-render one demo asset</summary>

```r
source("vignettes/helpers/render_algorithm_demo_assets.R")
render_algorithm_demo_assets(demos = "sweep_line")
```

Or from the shell:

```sh
R -q -f vignettes/helpers/render_algorithm_demo_assets.R --args demo=sweep_line
```

</details>

## Sweep line

A vertical line moves left-to-right across a collection of 1D intervals. At
each position, the algorithm asks: which intervals are active right now? This
is the canonical setting for a sweep over ordered events.

Events (interval start and end points) live in an `ordered_sequence` keyed by
x-coordinate; `pop_all_key()` pulls every event at the current sweep position
as a single batch. A second `ordered_sequence` — keyed by interval end — is
the evolving active set: start events insert, end events remove by key. Each
updated active set is pushed into a `flexseq` snapshot timeline. Because every
snapshot is itself a persistent value, the animation is just playback over
that timeline — no re-running, no state to rewind.

```{r}
items  <- c("a", "b", "c", "d")
starts <- c(1,    2,   3,   4)
ends   <- c(3,    4,   5,   6)

xs <- c(starts, ends) |> sort() |> unique()
first_x <- xs[1]

index <- as_interval_index(items, start = starts, end = ends)

active <- interval_index()
playback <- flexseq()

for(x in xs) {
  new_popped <- pop_all_point(index, x)
  active <- merge(active, new_popped$elements)
  index <- new_popped$remaining

  remove_popped <- pop_all_within(active, start = first_x, end = x, bounds = "()")
  active <- remove_popped$remaining
  
  cat("Currently active 2: ", unlist(active), "\n")
  
  playback <- push_back(playback, active)
}
```

```{r sweep-line-viz, echo=FALSE, fig.cap="Sweep-line interval activity over persistent snapshots."}
knitr::include_graphics(algorithm_demo_asset("sweep_line"))
```

## Segment sweep

A 2D variant: the sweep line crosses a set of line segments, and the demo
discovers pairwise intersections among segments that are simultaneously
active. It's a simplified version of the Bentley–Ottmann setup.

The event stream (segment start/end events, intersection events) is an
`ordered_sequence` keyed by x. Active segment candidates are queried from an
`interval_index` built on segment x-ranges. Interestingly, there is a *second*
`ordered_sequence` that accumulates discovered intersections as the sweep
progresses — a nice illustration that an ordered sequence is useful not only
as an event queue but as a growing sorted log. A `flexseq` holds the snapshot
timeline, so the animation can replay the sweep with the intersection list
growing step by step.

```{r segment-sweep-viz, echo=FALSE, fig.cap="Segment sweep with active-set intersection discovery."}
knitr::include_graphics(algorithm_demo_asset("segment_sweep"))
```

## Convex hull

Andrew's monotone chain: sort the points by `(x, y)`, sweep once to build the
lower hull, sweep again for the upper hull, and concatenate. Each sweep is a
stack algorithm — push the next point, then pop while the last three points
turn the wrong way.

This demo is notable because `flexseq` carries most of the work. The sorted
input lives in an `ordered_sequence`, but the lower and upper hull stacks are
themselves `flexseq`s, and every push and pop returns a new one rather than
mutating in place. A third `flexseq` stores the snapshot timeline, so after
the run you have a complete record of every intermediate hull — which point
was pushed, which points were popped because they violated convexity, and
what the partial hull looked like at every step.

```{r convex-hull-viz, echo=FALSE, fig.cap="Monotone-chain hull construction with persistent snapshots."}
knitr::include_graphics(algorithm_demo_asset("convex_hull"))
```

## Fortune frontier

Fortune's sweep-line algorithm for Voronoi diagrams. A vertical sweep moves
across a set of sites; each new site introduces a parabola to the "beach
line" frontier, and the breakpoints between adjacent parabolas trace out the
Voronoi edges as the sweep advances.

Three structures share the work. A `priority_queue` holds site events keyed
by x; `pop_min()` produces the next event due. The active sites along the
frontier live in an `ordered_sequence`, keyed by a composite score so they
stay in frontier order. A `flexseq` accumulates the *breakpoint trail* — every
Voronoi edge segment traced out so far — and a second `flexseq` holds the
snapshot timeline. The trail is an especially good fit for persistence: it
only ever grows, and the accumulated value from any past step is still
available for comparison or partial rendering with no bookkeeping.

```{r fortune-viz, echo=FALSE, fig.cap="Fortune-style frontier sweep and revealed Voronoi structure."}
knitr::include_graphics(algorithm_demo_asset("fortune_frontier"))
```

## A* search

Standard A* pathfinding on a generated maze, with the frontier ordered by
`f = g + h`. This is the demo where persistence pays off in a way that goes
beyond replay.

A `priority_queue` keyed by f-score holds the open set; `pop_min()` picks the
next node to expand. A `closed` `ordered_sequence` keyed by node index answers
the "already visited?" check via `peek_key()` in log time. A `flexseq`
snapshot timeline captures open/closed/current state at every expansion step.
Because the full search history is preserved as an immutable timeline, a
post-hoc aggregate like *pressure* — how many times each cell appeared on the
frontier over the course of the search — can be computed by folding over the
snapshots after the fact, rather than sprinkling counters through the search
loop. The large-maze animation exploits this to reveal the final path on top
of a pressure heatmap once the search settles.

```{r astar-viz, echo=FALSE, fig.cap="A* replay with staged final-route reveal, then persistent pressure summary."}
knitr::include_graphics(algorithm_demo_asset("astar_organic_history_large_fast"))
```

## Wave clear

A two-stage simulation over a weighted grid. Cells light up in Dijkstra-
distance order from a source, and a lagged "clear-wave" follows behind,
deactivating cells after a delay. The visible result is a travelling band.

Two `priority_queue`s drive the scheduling — one for add-events, one for
clear-events — both keyed by time; pop-by-priority-band retrieves everything
due at the current tick. The currently-active cell set lives in an
`ordered_sequence` keyed by cell index, with `insert()` on activation and
`pop_key()` on clear. A `flexseq` records the snapshot timeline. Because the
full temporal record is preserved, aggregate dynamics (peak concurrent active
count, first-arrival time at a goal cell) can be read off the history without
re-simulating.

```{r wave-clear-viz, echo=FALSE, fig.cap="Wave and clear-wave progression over persistent state snapshots."}
knitr::include_graphics(algorithm_demo_asset("wave_clear"))
```

## Source and rendering

The full implementations live under `vignettes/helpers/demos/`:

- `demo_sweep_line_immutables.R`
- `demo_segment_sweep_immutables.R`
- `demo_convex_hull_immutables.R`
- `demo_fortune_frontier_immutables.R`
- `demo_astar_immutables.R`
- `demo_wave_clear_immutables.R`

The vignette-side helpers (asset paths, render presets, the A* preset used by
the large-maze animation) are in
`vignettes/helpers/algorithm_demos_helpers.R`, and
`vignettes/helpers/render_algorithm_demo_assets.R` is the entry point for
re-rendering any or all of the GIFs.
