---
title: "Benchmarking immutables Collections"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Benchmarking immutables Collections}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
if(requireNamespace("pkgload", quietly = TRUE)) {
  pkgload::load_all(".", quiet = TRUE)
} else if(requireNamespace("immutables", quietly = TRUE)) {
  library(immutables)
} else {
  stop("Need either installed 'immutables' or the 'pkgload' package to render this vignette.")
}
source(file.path("helpers", "benchmark_helpers.R"), local = TRUE)
run_benchmarks <- benchmark_should_run()
```

## Method

This article measures individual operation times for each collection type
provided by `immutables`, following the operation-by-operation style of
Barry (2018). Each timing cell builds a fresh data structure of size *n*,
then uses `microbenchmark` to time one operation. We report the median of
`r benchmark_default_repeats()` repetitions per cell.

Comparisons use idiomatic base R implementations as controls:

- **Sequences and queues**: base R lists (plus `rstackdeque` for queues).
- **Priority queues**: parallel value and priority vectors with
  `which.min()`/`which.max()`.
- **Interval queries**: a `data.frame` with start/end columns, filtered
  with vectorized comparisons. When the Bioconductor package `IRanges` is
  available, it is included as a third comparison.

All `immutables` structures are persistent — operations return new
structures without modifying the input. The base R baselines are
copy-on-write but not persistent in the same sense; we benchmark the
cost of producing the updated copy.

```{r benchmark-note, eval = !run_benchmarks}
cat(
  "Live benchmark chunks are skipped by default.",
  "Set IMMUTABLES_RUN_BENCHMARK_VIGNETTE=1 before rendering",
  "to run the full grid."
)
```

```{r run-benchmarks, eval = run_benchmarks, cache = FALSE}
bench_results <- run_benchmark_vignette()
```

## Sequence operations

Seven operations on sequences of 100 to 500,000 elements. Append,
prepend, concatenate, and split exploit the finger tree's O(log n)
structural sharing; get, replace, and remove at the middle exercise
index-based splitting.

```{r sequence-plot, eval = run_benchmarks, fig.width = 9, fig.height = 7}
bench_results$sequence_plot
```

## Queue operations

FIFO enqueue (push to back) and dequeue (pop from front) compared across
`flexseq`, `rstackdeque`, and base R lists. Both `flexseq` and
`rstackdeque` provide O(log n) or amortized O(1) queue operations; base
R's `c()` and `[-1]` are O(n).

```{r queue-plot, eval = run_benchmarks, fig.width = 8, fig.height = 4.8}
bench_results$queue_plot
```

## Priority queue operations

Insert, peek, and pop operations for min and max priority. The base R
baseline stores values and priorities as parallel vectors and uses
`which.min()` / `which.max()` for queries — O(n) per call. The
`priority_queue` from `immutables` maintains a sorted finger tree with
cached min/max monoids, giving O(log n) insert and pop.

```{r pq-plot, eval = run_benchmarks, fig.width = 9, fig.height = 7}
bench_results$pq_plot
```

## Interval queries

Insert and three query types (single-point lookup, all-point matches,
and range overlap) on a collection of intervals with integer endpoints.
The base R baseline stores intervals in a `data.frame` and filters with
vectorized comparisons — simple and fast for small *n*, but O(n) per
query. `interval_index` uses an augmented finger tree for O(log n + k)
queries, where *k* is the number of matches.

```{r ivx-plot, eval = run_benchmarks, fig.width = 9, fig.height = 6}
bench_results$ivx_plot
```

## Notes

- All `immutables` operations are persistent: the original structure is
  never modified. Base R list operations produce shallow copies via
  copy-on-write.
- Sequence **append** and **prepend** are O(log n) for `flexseq` vs O(n)
  for base R; the crossover is visible at moderate sizes.
- **Get middle** (`[[`) is O(log n) for `flexseq` vs O(1) for base R —
  `flexseq` will always be slower here. This is the cost of tree-based
  indexing.
- Queue and priority queue operations show the expected O(log n) vs O(n)
  separation at larger sizes.
- Interval query performance depends on the number of matches (*k*). The
  benchmark uses a fixed query point and range, so *k* varies with *n*
  and the random interval layout.
- These figures are meant for article exposition rather than internal
  regression tracking.
