---
title: "Benchmarking immutables Collections"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Benchmarking immutables Collections}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")

loaded <- tryCatch({
  pkgload::load_all(quiet = TRUE)
  TRUE
}, error = function(e) FALSE)

if(!loaded) {
  library(immutables)
}

library(microbenchmark)
library(ggplot2)

run_benchmarks <- identical(Sys.getenv("IMMUTABLES_RUN_BENCHMARK_VIGNETTE"), "1")
repeats <- 8L

append_benchmark_rows <- function(rows, impl, op, n, time_us) {
  push_front(rows, data.frame(
    impl = impl,
    op = op,
    n = n,
    time_us = time_us
  ))
}

run_fresh_benchmark <- function(repeats, setup, bench) {
  gc(FALSE)
  time_us <- numeric(repeats)

  for(rep_idx in seq_len(repeats)) {
    state <- setup()
    benchmark_run <- local({
      microbenchmark(bench(state), times = 1L)
    })
    time_us[[rep_idx]] <- benchmark_run$time[[1L]] / 1000
  }

  time_us
}
```


## Method

This article measures individual operation times for each collection type
provided by `immutables`. Each recorded repetition rebuilds the fixture for
size *n*, then uses `microbenchmark(..., times = 1)` to time one operation on
that fresh state.

Comparisons use base R implementations and select structures from
`rstackdeque` and `IRanges`:

- **Sequences and queue operations**:
  - immutables: `flexseq()`
  - base R: `list()`
  - `rstackdeque`: `rpqueue()`
- **Priority operations**:
  - immutables: `priority_queue()`
  - base R: parallel value and priority vectors with
    `which.min()` / `which.max()`
- **Interval operations**:
  - immutables: `interval_index()`
  - base R: `data.frame` with start/end columns, filtered with vectorized
    comparisons
  - `IRanges`: `IRanges()`



## Sequence operations

Seven operations on sequences of 1,024 to 131,072 elements. Append, prepend,
concatenate, and split exploit the finger tree's O(log n) structural sharing;
get, replace, and remove at the middle exercise index-based splitting.

```{r sequence-bench, eval=run_benchmarks, cache=FALSE}
repeats <- 6L

sequence_sizes <- 2^(10 + 0:6)
sequence_rows <- flexseq()

for(sequence_size in sequence_sizes) {
  cat("Sequence ops, size ", sequence_size, "\n")
  make_sequence_values <- function() {
    as.list(sprintf("v_%06d", seq_len(sequence_size)))
  }
  middle_index <- as.integer(sequence_size / 2)

  time_us <- run_fresh_benchmark(
    repeats,
    setup = function() list(flex_sequence = as_flexseq(make_sequence_values())),
    bench = function(state) push_back(state$flex_sequence, "z")
  )
  sequence_rows <- append_benchmark_rows(sequence_rows, "flexseq", "append", sequence_size, time_us)

  time_us <- run_fresh_benchmark(
    repeats,
    setup = function() list(flex_sequence = as_flexseq(make_sequence_values())),
    bench = function(state) push_front(state$flex_sequence, "z")
  )
  sequence_rows <- append_benchmark_rows(sequence_rows, "flexseq", "prepend", sequence_size, time_us)

  time_us <- run_fresh_benchmark(
    repeats,
    setup = function() list(flex_sequence = as_flexseq(make_sequence_values()), middle_index = middle_index),
    bench = function(state) state$flex_sequence[[state$middle_index]]
  )
  sequence_rows <- append_benchmark_rows(sequence_rows, "flexseq", "get middle", sequence_size, time_us)

  time_us <- run_fresh_benchmark(
    repeats,
    setup = function() list(flex_sequence = as_flexseq(make_sequence_values()), middle_index = middle_index),
    bench = function(state) {
      updated_sequence <- state$flex_sequence
      updated_sequence[[state$middle_index]] <- "y"
      updated_sequence
    }
  )
  sequence_rows <- append_benchmark_rows(sequence_rows, "flexseq", "replace middle", sequence_size, time_us)

  time_us <- run_fresh_benchmark(
    repeats,
    setup = function() list(flex_sequence = as_flexseq(make_sequence_values()), middle_index = middle_index),
    bench = function(state) pop_at(state$flex_sequence, state$middle_index)$remaining
  )
  sequence_rows <- append_benchmark_rows(sequence_rows, "flexseq", "remove middle", sequence_size, time_us)

  time_us <- run_fresh_benchmark(
    repeats,
    setup = function() list(
      left = as_flexseq(make_sequence_values()),
      right = as_flexseq(make_sequence_values())
    ),
    bench = function(state) c(state$left, state$right)
  )
  sequence_rows <- append_benchmark_rows(sequence_rows, "flexseq", "concatenate", sequence_size, time_us)

  time_us <- run_fresh_benchmark(
    repeats,
    setup = function() list(flex_sequence = as_flexseq(make_sequence_values()), middle_index = middle_index),
    bench = function(state) split_at(state$flex_sequence, state$middle_index)
  )
  sequence_rows <- append_benchmark_rows(sequence_rows, "flexseq", "split at middle", sequence_size, time_us)

  time_us <- run_fresh_benchmark(
    repeats,
    setup = function() list(base_list_sequence = make_sequence_values()),
    bench = function(state) c(state$base_list_sequence, list("z"))
  )
  sequence_rows <- append_benchmark_rows(sequence_rows, "base R list", "append", sequence_size, time_us)

  time_us <- run_fresh_benchmark(
    repeats,
    setup = function() list(base_list_sequence = make_sequence_values()),
    bench = function(state) c(list("z"), state$base_list_sequence)
  )
  sequence_rows <- append_benchmark_rows(sequence_rows, "base R list", "prepend", sequence_size, time_us)

  time_us <- run_fresh_benchmark(
    repeats,
    setup = function() list(base_list_sequence = make_sequence_values(), middle_index = middle_index),
    bench = function(state) state$base_list_sequence[[state$middle_index]]
  )
  sequence_rows <- append_benchmark_rows(sequence_rows, "base R list", "get middle", sequence_size, time_us)

  time_us <- run_fresh_benchmark(
    repeats,
    setup = function() list(base_list_sequence = make_sequence_values(), middle_index = middle_index),
    bench = function(state) {
      updated_list_sequence <- state$base_list_sequence
      updated_list_sequence[[state$middle_index]] <- "y"
      updated_list_sequence
    }
  )
  sequence_rows <- append_benchmark_rows(sequence_rows, "base R list", "replace middle", sequence_size, time_us)

  time_us <- run_fresh_benchmark(
    repeats,
    setup = function() list(base_list_sequence = make_sequence_values(), middle_index = middle_index),
    bench = function(state) state$base_list_sequence[-state$middle_index]
  )
  sequence_rows <- append_benchmark_rows(sequence_rows, "base R list", "remove middle", sequence_size, time_us)

  time_us <- run_fresh_benchmark(
    repeats,
    setup = function() list(
      left = make_sequence_values(),
      right = make_sequence_values()
    ),
    bench = function(state) c(state$left, state$right)
  )
  sequence_rows <- append_benchmark_rows(sequence_rows, "base R list", "concatenate", sequence_size, time_us)

  time_us <- run_fresh_benchmark(
    repeats,
    setup = function() list(base_list_sequence = make_sequence_values(), middle_index = middle_index),
    bench = function(state) {
      list(
        left = state$base_list_sequence[seq_len(state$middle_index - 1L)],
        value = state$base_list_sequence[[state$middle_index]],
        right = state$base_list_sequence[(state$middle_index + 1L):sequence_size]
      )
    }
  )
  sequence_rows <- append_benchmark_rows(sequence_rows, "base R list", "split at middle", sequence_size, time_us)
}

seq_results <- do.call(rbind, as.list(sequence_rows))
```


```{r sequence-plot, eval=run_benchmarks, fig.width=9, fig.height=7}
seq_all <- seq_results
seq_all$time_ms <- seq_all$time_us / 1000

sorted_sizes <- sort(unique(seq_all$n))
pow_labels <- lapply(sorted_sizes, function(s) bquote(2^.(log2(s)))) |> as.character()
seq_all$n_cat <- factor(seq_all$n, levels = sorted_sizes)

ggplot(seq_all, aes(x = n_cat, y = time_ms, color = impl)) +
  geom_point(position = position_jitter(width = 0.15, height = 0)) +
  facet_wrap(~ op, scales = "free_y") +
  scale_x_discrete(labels = pow_labels) +
  labs(
    title = "Sequence Operations",
    x = "Number of elements",
    y = "Time (ms)",
    color = "Implementation"
  ) +
  theme_bw() +
  theme(plot.title = element_text(hjust = 0.5), legend.position = "bottom")
```


## Queue operations

FIFO enqueue (push to back) and dequeue (pop from front) compared across
`flexseq`, `rstackdeque`, and base R lists. Both `flexseq` and
`rstackdeque` provide O(log n) or amortized O(1) queue operations; base R's
`c()` and `[-1]` are O(n).

```{r queue-bench, eval=run_benchmarks, cache=FALSE}
queue_sizes <- 2^(10 + 0:7)
queue_rows <- flexseq()

for(queue_size in queue_sizes) {
  make_queue_items <- function() {
    as.list(rep("queue_item", queue_size))
  }

  time_us <- run_fresh_benchmark(
    repeats,
    setup = function() list(flex_queue = as_flexseq(make_queue_items())),
    bench = function(state) push_back(state$flex_queue, "d")
  )
  queue_rows <- append_benchmark_rows(queue_rows, "flexseq", "enqueue", queue_size, time_us)

  time_us <- run_fresh_benchmark(
    repeats,
    setup = function() list(flex_queue = as_flexseq(make_queue_items())),
    bench = function(state) pop_front(state$flex_queue)$remaining
  )
  queue_rows <- append_benchmark_rows(queue_rows, "flexseq", "dequeue", queue_size, time_us)

  time_us <- run_fresh_benchmark(
    repeats,
    setup = function() list(rstackdeque_queue = rstackdeque::as.rpqueue(make_queue_items())),
    bench = function(state) rstackdeque::insert_back(state$rstackdeque_queue, "d")
  )
  queue_rows <- append_benchmark_rows(queue_rows, "rstackdeque", "enqueue", queue_size, time_us)

  time_us <- run_fresh_benchmark(
    repeats,
    setup = function() list(rstackdeque_queue = rstackdeque::as.rpqueue(make_queue_items())),
    bench = function(state) rstackdeque::without_front(state$rstackdeque_queue)
  )
  queue_rows <- append_benchmark_rows(queue_rows, "rstackdeque", "dequeue", queue_size, time_us)

  time_us <- run_fresh_benchmark(
    repeats,
    setup = function() list(base_list_queue = make_queue_items()),
    bench = function(state) c(state$base_list_queue, list("d"))
  )
  queue_rows <- append_benchmark_rows(queue_rows, "base R list", "enqueue", queue_size, time_us)

  time_us <- run_fresh_benchmark(
    repeats,
    setup = function() list(base_list_queue = make_queue_items()),
    bench = function(state) state$base_list_queue[-1L]
  )
  queue_rows <- append_benchmark_rows(queue_rows, "base R list", "dequeue", queue_size, time_us)
}

queue_results <- do.call(rbind, as.list(queue_rows))
```
```{r queue-plot, eval=run_benchmarks, fig.width=8, fig.height=4.8}
queue_all <- queue_results
queue_all$time_ms <- queue_all$time_us / 1000
queue_all$n_cat <- factor(queue_all$n, levels = sort(unique(queue_all$n)))

ggplot(queue_all, aes(x = n_cat, y = time_ms, color = impl)) +
  geom_boxplot() +
  facet_wrap(~ op, scales = "free_y") +
  labs(
    title = "Queue Operations",
    x = "Number of elements",
    y = "Time (ms)",
    color = "Implementation"
  ) +
  theme_bw() +
  theme(plot.title = element_text(hjust = 0.5), legend.position = "bottom")
```


## Priority queue operations

Insert, peek, and pop operations for min and max priority. The base R
baseline stores values and priorities as parallel vectors and uses
`which.min()` / `which.max()` for queries, so those reads are O(n) per call.
`priority_queue` maintains a sorted finger tree with cached min/max monoids,
giving O(log n) insert and pop.

```{r pq-bench, eval=run_benchmarks, cache=FALSE}
priority_queue_sizes <- c(100, 500, 1000, 5000, 10000, 50000)
priority_queue_rows <- flexseq()

set.seed(42)
max_priority_queue_size <- max(priority_queue_sizes)
all_priority_values <- sprintf("val_%06d", seq_len(max_priority_queue_size))
all_priority_weights <- runif(max_priority_queue_size)

for(priority_queue_size in priority_queue_sizes) {
  priority_values <- as.list(all_priority_values[seq_len(priority_queue_size)])
  priority_weights <- all_priority_weights[seq_len(priority_queue_size)]

  time_us <- run_fresh_benchmark(
    repeats,
    setup = function() list(
      priority_queue_fixture = as_priority_queue(priority_values, priorities = priority_weights)
    ),
    bench = function(state) insert(state$priority_queue_fixture, "val_new", 0.5)
  )
  priority_queue_rows <- append_benchmark_rows(priority_queue_rows, "priority_queue", "insert", priority_queue_size, time_us)

  time_us <- run_fresh_benchmark(
    repeats,
    setup = function() list(
      priority_queue_fixture = as_priority_queue(priority_values, priorities = priority_weights)
    ),
    bench = function(state) peek_min(state$priority_queue_fixture)
  )
  priority_queue_rows <- append_benchmark_rows(priority_queue_rows, "priority_queue", "peek min", priority_queue_size, time_us)

  time_us <- run_fresh_benchmark(
    repeats,
    setup = function() list(
      priority_queue_fixture = as_priority_queue(priority_values, priorities = priority_weights)
    ),
    bench = function(state) pop_min(state$priority_queue_fixture)$remaining
  )
  priority_queue_rows <- append_benchmark_rows(priority_queue_rows, "priority_queue", "pop min", priority_queue_size, time_us)

  time_us <- run_fresh_benchmark(
    repeats,
    setup = function() list(
      priority_queue_fixture = as_priority_queue(priority_values, priorities = priority_weights)
    ),
    bench = function(state) peek_max(state$priority_queue_fixture)
  )
  priority_queue_rows <- append_benchmark_rows(priority_queue_rows, "priority_queue", "peek max", priority_queue_size, time_us)

  time_us <- run_fresh_benchmark(
    repeats,
    setup = function() list(
      priority_queue_fixture = as_priority_queue(priority_values, priorities = priority_weights)
    ),
    bench = function(state) pop_max(state$priority_queue_fixture)$remaining
  )
  priority_queue_rows <- append_benchmark_rows(priority_queue_rows, "priority_queue", "pop max", priority_queue_size, time_us)

  time_us <- run_fresh_benchmark(
    repeats,
    setup = function() list(
      base_priority_queue = list(
        values = all_priority_values[seq_len(priority_queue_size)],
        priorities = priority_weights
      )
    ),
    bench = function(state) {
      list(
        values = c(state$base_priority_queue$values, "val_new"),
        priorities = c(state$base_priority_queue$priorities, 0.5)
      )
    }
  )
  priority_queue_rows <- append_benchmark_rows(priority_queue_rows, "base R", "insert", priority_queue_size, time_us)

  time_us <- run_fresh_benchmark(
    repeats,
    setup = function() list(
      base_priority_queue = list(
        values = all_priority_values[seq_len(priority_queue_size)],
        priorities = priority_weights
      )
    ),
    bench = function(state) {
      state$base_priority_queue$values[which.min(state$base_priority_queue$priorities)]
    }
  )
  priority_queue_rows <- append_benchmark_rows(priority_queue_rows, "base R", "peek min", priority_queue_size, time_us)

  time_us <- run_fresh_benchmark(
    repeats,
    setup = function() list(
      base_priority_queue = list(
        values = all_priority_values[seq_len(priority_queue_size)],
        priorities = priority_weights
      )
    ),
    bench = function(state) {
      min_index <- which.min(state$base_priority_queue$priorities)
      list(
        values = state$base_priority_queue$values[-min_index],
        priorities = state$base_priority_queue$priorities[-min_index]
      )
    }
  )
  priority_queue_rows <- append_benchmark_rows(priority_queue_rows, "base R", "pop min", priority_queue_size, time_us)

  time_us <- run_fresh_benchmark(
    repeats,
    setup = function() list(
      base_priority_queue = list(
        values = all_priority_values[seq_len(priority_queue_size)],
        priorities = priority_weights
      )
    ),
    bench = function(state) {
      state$base_priority_queue$values[which.max(state$base_priority_queue$priorities)]
    }
  )
  priority_queue_rows <- append_benchmark_rows(priority_queue_rows, "base R", "peek max", priority_queue_size, time_us)

  time_us <- run_fresh_benchmark(
    repeats,
    setup = function() list(
      base_priority_queue = list(
        values = all_priority_values[seq_len(priority_queue_size)],
        priorities = priority_weights
      )
    ),
    bench = function(state) {
      max_index <- which.max(state$base_priority_queue$priorities)
      list(
        values = state$base_priority_queue$values[-max_index],
        priorities = state$base_priority_queue$priorities[-max_index]
      )
    }
  )
  priority_queue_rows <- append_benchmark_rows(priority_queue_rows, "base R", "pop max", priority_queue_size, time_us)
}

pq_results <- do.call(rbind, as.list(priority_queue_rows))
```


```{r pq-plot, eval=run_benchmarks, fig.width=9, fig.height=7}
pq_all <- pq_results
pq_all$time_ms <- pq_all$time_us / 1000
pq_medians <- aggregate(time_us ~ impl + op + n, data = pq_results, FUN = median)
pq_medians$time_ms <- pq_medians$time_us / 1000

ggplot(pq_all, aes(x = n, y = time_ms, color = impl)) +
  geom_point(alpha = 0.25, size = 1.2, position = position_jitter(width = 0.03)) +
  geom_line(data = pq_medians, linewidth = 0.6) +
  geom_point(data = pq_medians, size = 1.8) +
  facet_wrap(~ op, scales = "free_y") +
  scale_x_log10(labels = scales::label_comma()) +
  scale_y_log10(labels = scales::label_comma()) +
  labs(
    title = "Priority Queue Operations",
    x = "Number of elements",
    y = "Time (ms)",
    color = "Implementation"
  ) +
  theme_bw() +
  theme(plot.title = element_text(hjust = 0.5), legend.position = "bottom")
```


## Interval queries

Insert and three query types (single-point lookup, all-point matches, and
range overlap) on a collection of intervals with integer endpoints. The base R
baseline stores intervals in a `data.frame` and filters with vectorized
comparisons, which is simple and fast for small *n* but O(n) per query.
`interval_index` uses an augmented finger tree for O(log n + k) queries, where
*k* is the number of matches.

```{r ivx-bench, eval=run_benchmarks, cache=FALSE}
interval_index_sizes <- c(100, 500, 1000, 5000, 10000, 50000)
interval_index_rows <- flexseq()

set.seed(123)
max_interval_index_size <- max(interval_index_sizes)
all_interval_starts <- sort(sample.int(max_interval_index_size * 10L, max_interval_index_size))
all_interval_widths <- sample.int(100L, max_interval_index_size, replace = TRUE)
all_interval_ends <- all_interval_starts + all_interval_widths
all_interval_values <- sprintf("interval_%06d", seq_len(max_interval_index_size))

query_point <- all_interval_starts[as.integer(max_interval_index_size / 2)] + 10L
query_lower_bound <- all_interval_starts[as.integer(max_interval_index_size * 0.4)]
query_upper_bound <- all_interval_starts[as.integer(max_interval_index_size * 0.5)]

has_iranges <- requireNamespace("IRanges", quietly = TRUE) &&
  requireNamespace("S4Vectors", quietly = TRUE)

for(interval_index_size in interval_index_sizes) {
  interval_starts <- all_interval_starts[seq_len(interval_index_size)]
  interval_ends <- all_interval_ends[seq_len(interval_index_size)]
  interval_values <- all_interval_values[seq_len(interval_index_size)]
  inserted_interval_start <- all_interval_starts[as.integer(max_interval_index_size / 2)]
  inserted_interval_end <- inserted_interval_start + 50L

  time_us <- run_fresh_benchmark(
    repeats,
    setup = function() list(
      interval_index_fixture = as_interval_index(
        as.list(interval_values),
        start = interval_starts,
        end = interval_ends,
        bounds = "[]"
      )
    ),
    bench = function(state) {
      insert(state$interval_index_fixture, "interval_new", inserted_interval_start, inserted_interval_end)
    }
  )
  interval_index_rows <- append_benchmark_rows(interval_index_rows, "interval_index", "insert", interval_index_size, time_us)

  time_us <- run_fresh_benchmark(
    repeats,
    setup = function() list(
      interval_index_fixture = as_interval_index(
        as.list(interval_values),
        start = interval_starts,
        end = interval_ends,
        bounds = "[]"
      )
    ),
    bench = function(state) peek_point(state$interval_index_fixture, query_point, bounds = "[]")
  )
  interval_index_rows <- append_benchmark_rows(interval_index_rows, "interval_index", "point query", interval_index_size, time_us)

  time_us <- run_fresh_benchmark(
    repeats,
    setup = function() list(
      interval_index_fixture = as_interval_index(
        as.list(interval_values),
        start = interval_starts,
        end = interval_ends,
        bounds = "[]"
      )
    ),
    bench = function(state) peek_all_point(state$interval_index_fixture, query_point, bounds = "[]")
  )
  interval_index_rows <- append_benchmark_rows(interval_index_rows, "interval_index", "all point matches", interval_index_size, time_us)

  time_us <- run_fresh_benchmark(
    repeats,
    setup = function() list(
      interval_index_fixture = as_interval_index(
        as.list(interval_values),
        start = interval_starts,
        end = interval_ends,
        bounds = "[]"
      )
    ),
    bench = function(state) {
      peek_all_overlaps(state$interval_index_fixture, query_lower_bound, query_upper_bound, bounds = "[]")
    }
  )
  interval_index_rows <- append_benchmark_rows(interval_index_rows, "interval_index", "overlap query", interval_index_size, time_us)

  time_us <- run_fresh_benchmark(
    repeats,
    setup = function() list(
      base_interval_data = data.frame(
        start = interval_starts,
        end = interval_ends,
        value = interval_values,
        stringsAsFactors = FALSE
      )
    ),
    bench = function(state) {
      rbind(
        state$base_interval_data,
        data.frame(
          start = inserted_interval_start,
          end = inserted_interval_end,
          value = "interval_new",
          stringsAsFactors = FALSE
        )
      )
    }
  )
  interval_index_rows <- append_benchmark_rows(interval_index_rows, "base R", "insert", interval_index_size, time_us)

  time_us <- run_fresh_benchmark(
    repeats,
    setup = function() list(
      base_interval_data = data.frame(
        start = interval_starts,
        end = interval_ends,
        value = interval_values,
        stringsAsFactors = FALSE
      )
    ),
    bench = function(state) {
      matching_rows <- which(state$base_interval_data$start <= query_point & query_point <= state$base_interval_data$end)
      if(length(matching_rows)) state$base_interval_data$value[matching_rows[1L]] else NULL
    }
  )
  interval_index_rows <- append_benchmark_rows(interval_index_rows, "base R", "point query", interval_index_size, time_us)

  time_us <- run_fresh_benchmark(
    repeats,
    setup = function() list(
      base_interval_data = data.frame(
        start = interval_starts,
        end = interval_ends,
        value = interval_values,
        stringsAsFactors = FALSE
      )
    ),
    bench = function(state) {
      state$base_interval_data[
        state$base_interval_data$start <= query_point & query_point <= state$base_interval_data$end,
        ,
        drop = FALSE
      ]
    }
  )
  interval_index_rows <- append_benchmark_rows(interval_index_rows, "base R", "all point matches", interval_index_size, time_us)

  time_us <- run_fresh_benchmark(
    repeats,
    setup = function() list(
      base_interval_data = data.frame(
        start = interval_starts,
        end = interval_ends,
        value = interval_values,
        stringsAsFactors = FALSE
      )
    ),
    bench = function(state) {
      state$base_interval_data[
        state$base_interval_data$start <= query_upper_bound &
        state$base_interval_data$end >= query_lower_bound,
        ,
        drop = FALSE
      ]
    }
  )
  interval_index_rows <- append_benchmark_rows(interval_index_rows, "base R", "overlap query", interval_index_size, time_us)

  if(has_iranges) {
    time_us <- run_fresh_benchmark(
      repeats,
      setup = function() list(
        iranges_fixture = list(
          iranges = IRanges::IRanges(start = interval_starts, end = interval_ends),
          values = interval_values
        )
      ),
      bench = function(state) {
        list(
          iranges = c(
            state$iranges_fixture$iranges,
            IRanges::IRanges(start = inserted_interval_start, end = inserted_interval_end)
          ),
          values = c(state$iranges_fixture$values, "interval_new")
        )
      }
    )
    interval_index_rows <- append_benchmark_rows(interval_index_rows, "IRanges", "insert", interval_index_size, time_us)

    time_us <- run_fresh_benchmark(
      repeats,
      setup = function() list(
        iranges_fixture = list(
          iranges = IRanges::IRanges(start = interval_starts, end = interval_ends),
          values = interval_values
        )
      ),
      bench = function(state) {
        point_query_range <- IRanges::IRanges(start = query_point, width = 1L)
        overlap_hits <- IRanges::findOverlaps(point_query_range, state$iranges_fixture$iranges)
        subject_hit_indices <- S4Vectors::subjectHits(overlap_hits)
        if(length(subject_hit_indices)) state$iranges_fixture$values[subject_hit_indices[1L]] else NULL
      }
    )
    interval_index_rows <- append_benchmark_rows(interval_index_rows, "IRanges", "point query", interval_index_size, time_us)

    time_us <- run_fresh_benchmark(
      repeats,
      setup = function() list(
        iranges_fixture = list(
          iranges = IRanges::IRanges(start = interval_starts, end = interval_ends),
          values = interval_values
        )
      ),
      bench = function(state) {
        point_query_range <- IRanges::IRanges(start = query_point, width = 1L)
        overlap_hits <- IRanges::findOverlaps(point_query_range, state$iranges_fixture$iranges)
        state$iranges_fixture$values[S4Vectors::subjectHits(overlap_hits)]
      }
    )
    interval_index_rows <- append_benchmark_rows(interval_index_rows, "IRanges", "all point matches", interval_index_size, time_us)

    time_us <- run_fresh_benchmark(
      repeats,
      setup = function() list(
        iranges_fixture = list(
          iranges = IRanges::IRanges(start = interval_starts, end = interval_ends),
          values = interval_values
        )
      ),
      bench = function(state) {
        overlap_query_range <- IRanges::IRanges(start = query_lower_bound, end = query_upper_bound)
        overlap_hits <- IRanges::findOverlaps(overlap_query_range, state$iranges_fixture$iranges)
        state$iranges_fixture$values[S4Vectors::subjectHits(overlap_hits)]
      }
    )
    interval_index_rows <- append_benchmark_rows(interval_index_rows, "IRanges", "overlap query", interval_index_size, time_us)
  }
}

ivx_results <- do.call(rbind, as.list(interval_index_rows))
```
```{r ivx-plot, eval=run_benchmarks, fig.width=9, fig.height=6}
ivx_all <- ivx_results
ivx_all$time_ms <- ivx_all$time_us / 1000
ivx_medians <- aggregate(time_us ~ impl + op + n, data = ivx_results, FUN = median)
ivx_medians$time_ms <- ivx_medians$time_us / 1000

ggplot(ivx_all, aes(x = n, y = time_ms, color = impl)) +
  geom_point(alpha = 0.25, size = 1.2, position = position_jitter(width = 0.03)) +
  geom_line(data = ivx_medians, linewidth = 0.6) +
  geom_point(data = ivx_medians, size = 1.8) +
  facet_wrap(~ op, scales = "free_y") +
  scale_x_log10(labels = scales::label_comma()) +
  scale_y_log10(labels = scales::label_comma()) +
  labs(
    title = "Interval Index Queries",
    x = "Number of elements",
    y = "Time (ms)",
    color = "Implementation"
  ) +
  theme_bw() +
  theme(plot.title = element_text(hjust = 0.5), legend.position = "bottom")
```
