---
title: "Benchmarking immutables Collections"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Benchmarking immutables Collections}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")

loaded <- tryCatch({
  pkgload::load_all(quiet = TRUE)
  TRUE
}, error = function(e) FALSE)

if(!loaded) {
  library(immutables)
}

library(microbenchmark)
library(ggplot2)

# Two env-var gates control this vignette:
#   IMMUTABLES_RUN_SLOW=true       -> re-run all benchmarks and save
#                                     results to inst/extdata/benchmarks.rds
#   IMMUTABLES_SAVE_FIGURES=true   -> also write pub-ready figures to
#                                     paper/figures/*.pdf
# When neither is set (CRAN, ordinary pkgdown build) the vignette loads
# cached results and renders plots without re-timing.
run_benchmarks <- identical(Sys.getenv("IMMUTABLES_RUN_SLOW"), "true")
save_figures   <- identical(Sys.getenv("IMMUTABLES_SAVE_FIGURES"), "true")
repeats <- 8L

# Accumulator populated either by the benchmark chunks (when
# run_benchmarks) or by the loader chunk below (otherwise).
results_list <- list()

bench_one <- function(rows, impl, op, n, repeats, setup, bench) {
  gc(FALSE)
  time_us <- numeric(repeats)
  for(i in seq_len(repeats)) {
    state <- setup()
    gc(FALSE)
    time_us[[i]] <- microbenchmark(bench(state), times = 1L)$time[[1L]] / 1000
  }
  push_back(rows, data.frame(impl = impl, op = op, n = n, time_us = time_us))
}
```

```{r load-cached, eval=!run_benchmarks, include=FALSE}
.cache_path <- system.file("extdata", "benchmarks.rds", package = "immutables")
if(nzchar(.cache_path) && file.exists(.cache_path)) {
  results_list <- readRDS(.cache_path)
}
```

## Method

This article measures individual operation times for each collection type
provided by `immutables`. Each recorded repetition rebuilds the fixture for
size *n*, then uses `microbenchmark(..., times = 1)` to time one operation on
that fresh state.

The numbers shown below are loaded from cached results shipped with the
package; re-running `data-raw/generate_publication_results.R` regenerates
them. See the script's header for runtime and options.

Comparisons use base R implementations and select structures from
`rstackdeque` and `IRanges`:

- **Sequences and queue operations**:
  - immutables: `flexseq()`
  - base R: `list()`
  - `rstackdeque`: `rpqueue()`
- **Priority operations**:
  - immutables: `priority_queue()`
  - base R: parallel value and priority vectors with
    `which.min()` / `which.max()`
- **Interval operations**:
  - immutables: `interval_index()`
  - base R: `data.frame` with start/end columns, filtered with vectorized
    comparisons
  - `IRanges`: `IRanges()`



## Sequence operations

Seven operations on sequences of 1,024 to 131,072 elements. Append, prepend,
concatenate, and split exploit the finger tree's O(log n) structural sharing;
get, replace, and remove at the middle exercise index-based splitting.

```{r sequence-bench, eval=run_benchmarks, cache=FALSE}
repeats <- 6L

sequence_sizes <- 2^(10 + 0:6)
rows <- flexseq()

for(n in sequence_sizes) {
  cat("Sequence ops, size ", n, "\n")
  vals <- function() as.list(sprintf("v_%06d", seq_len(n)))
  mid <- as.integer(n / 2)
  flex_setup <- function() list(s = as_flexseq(vals()), mid = mid)
  list_setup <- function() list(s = vals(), mid = mid)
  pair_flex  <- function() list(a = as_flexseq(vals()), b = as_flexseq(vals()))
  pair_list  <- function() list(a = vals(), b = vals())

  rows <- bench_one(rows, "flexseq", "append",  n, repeats, flex_setup,
    function(st) push_back(st$s, "z"))
  rows <- bench_one(rows, "flexseq", "prepend", n, repeats, flex_setup,
    function(st) push_front(st$s, "z"))
  rows <- bench_one(rows, "flexseq", "get middle", n, repeats, flex_setup,
    function(st) st$s[[st$mid]])
  rows <- bench_one(rows, "flexseq", "replace middle", n, repeats, flex_setup,
    function(st) { s <- st$s; s[[st$mid]] <- "y"; s })
  rows <- bench_one(rows, "flexseq", "remove middle", n, repeats, flex_setup,
    function(st) pop_at(st$s, st$mid)$remaining)
  rows <- bench_one(rows, "flexseq", "concatenate", n, repeats, pair_flex,
    function(st) c(st$a, st$b))
  rows <- bench_one(rows, "flexseq", "split at middle", n, repeats, flex_setup,
    function(st) split_at(st$s, st$mid))

  rows <- bench_one(rows, "base R list", "append",  n, repeats, list_setup,
    function(st) c(st$s, list("z")))
  rows <- bench_one(rows, "base R list", "prepend", n, repeats, list_setup,
    function(st) c(list("z"), st$s))
  rows <- bench_one(rows, "base R list", "get middle", n, repeats, list_setup,
    function(st) st$s[[st$mid]])
  rows <- bench_one(rows, "base R list", "replace middle", n, repeats, list_setup,
    function(st) { s <- st$s; s[[st$mid]] <- "y"; s })
  rows <- bench_one(rows, "base R list", "remove middle", n, repeats, list_setup,
    function(st) st$s[-st$mid])
  rows <- bench_one(rows, "base R list", "concatenate", n, repeats, pair_list,
    function(st) c(st$a, st$b))
  rows <- bench_one(rows, "base R list", "split at middle", n, repeats, list_setup,
    function(st) list(
      left = st$s[seq_len(st$mid - 1L)],
      value = st$s[[st$mid]],
      right = st$s[(st$mid + 1L):n]
    ))
}

results_list$sequence <- do.call(rbind, as.list(rows))
```
```{r sequence-plot, fig.width=9, fig.height=7}
if(!is.null(results_list$sequence)) {
  seq_results <- results_list$sequence
  seq_results$time_ms <- seq_results$time_us / 1000
  sorted_sizes <- sort(unique(seq_results$n))
  pow_labels <- lapply(sorted_sizes, function(s) bquote(2^.(log2(s)))) |> as.character()
  seq_results$n_cat <- factor(seq_results$n, levels = sorted_sizes)

  p_sequence <- ggplot(seq_results, aes(x = n_cat, y = time_ms, color = impl)) +
    geom_point(position = position_jitter(width = 0.15, height = 0)) +
    facet_wrap(~ op, scales = "free_y") +
    scale_x_discrete(labels = pow_labels) +
    labs(
      title = "Sequence Operations",
      x = "Number of elements",
      y = "Time (ms)",
      color = "Implementation"
    ) +
    theme_bw() +
    theme(plot.title = element_text(hjust = 0.5), legend.position = "bottom")
  print(p_sequence)
} else {
  knitr::asis_output("*Benchmark results not yet generated. Run `data-raw/generate_publication_results.R` to populate.*")
}
```




## Queue operations

FIFO enqueue (push to back) and dequeue (pop from front) compared across
`flexseq`, `rstackdeque`, and base R lists. Both `flexseq` and
`rstackdeque` provide O(log n) or amortized O(1) queue operations; base R's
`c()` and `[-1]` are O(n).

```{r queue-bench, eval=run_benchmarks, cache=FALSE}
queue_sizes <- 2^(10 + 0:7)
rows <- flexseq()

for(n in queue_sizes) {
  items <- function() as.list(rep("queue_item", n))
  flex_setup <- function() list(q = as_flexseq(items()))
  rsd_setup  <- function() list(q = rstackdeque::as.rpqueue(items()))
  list_setup <- function() list(q = items())

  rows <- bench_one(rows, "flexseq",     "enqueue", n, repeats, flex_setup,
    function(st) push_back(st$q, "d"))
  rows <- bench_one(rows, "flexseq",     "dequeue", n, repeats, flex_setup,
    function(st) pop_front(st$q)$remaining)
  rows <- bench_one(rows, "rstackdeque", "enqueue", n, repeats, rsd_setup,
    function(st) rstackdeque::insert_back(st$q, "d"))
  rows <- bench_one(rows, "rstackdeque", "dequeue", n, repeats, rsd_setup,
    function(st) rstackdeque::without_front(st$q))
  rows <- bench_one(rows, "base R list", "enqueue", n, repeats, list_setup,
    function(st) c(st$q, list("d")))
  rows <- bench_one(rows, "base R list", "dequeue", n, repeats, list_setup,
    function(st) st$q[-1L])
}

results_list$queue <- do.call(rbind, as.list(rows))
```
```{r queue-plot, fig.width=8, fig.height=4.8}
if(!is.null(results_list$queue)) {
  queue_results <- results_list$queue
  queue_results$time_ms <- queue_results$time_us / 1000
  queue_results$n_cat <- factor(queue_results$n, levels = sort(unique(queue_results$n)))

  p_queue <- ggplot(queue_results, aes(x = n_cat, y = time_ms, color = impl)) +
    geom_boxplot() +
    facet_wrap(~ op, scales = "free_y") +
    labs(
      title = "Queue Operations",
      x = "Number of elements",
      y = "Time (ms)",
      color = "Implementation"
    ) +
    theme_bw() +
    theme(plot.title = element_text(hjust = 0.5), legend.position = "bottom")
  print(p_queue)
} else {
  knitr::asis_output("*Benchmark results not yet generated.*")
}
```


## Priority queue operations

Insert, peek, and pop operations for min and max priority. The base R
baseline stores values and priorities as parallel vectors and uses
`which.min()` / `which.max()` for queries, so those reads are O(n) per call.
`priority_queue` maintains a sorted finger tree with cached min/max monoids,
giving O(log n) insert and pop.

```{r pq-bench, eval=run_benchmarks, cache=FALSE}
pq_sizes <- c(100, 500, 1000, 5000, 10000, 50000)
rows <- flexseq()

set.seed(42)
max_pq <- max(pq_sizes)
all_pq_vals <- sprintf("val_%06d", seq_len(max_pq))
all_pq_pri  <- runif(max_pq)

for(n in pq_sizes) {
  pv <- as.list(all_pq_vals[seq_len(n)])
  pw <- all_pq_pri[seq_len(n)]
  pq_setup   <- function() list(pq = as_priority_queue(pv, priorities = pw))
  base_setup <- function() list(v = all_pq_vals[seq_len(n)], p = pw)

  rows <- bench_one(rows, "priority_queue", "insert",   n, repeats, pq_setup,
    function(st) insert(st$pq, "val_new", 0.5))
  rows <- bench_one(rows, "priority_queue", "peek min", n, repeats, pq_setup,
    function(st) peek_min(st$pq))
  rows <- bench_one(rows, "priority_queue", "pop min",  n, repeats, pq_setup,
    function(st) pop_min(st$pq)$remaining)
  rows <- bench_one(rows, "priority_queue", "peek max", n, repeats, pq_setup,
    function(st) peek_max(st$pq))
  rows <- bench_one(rows, "priority_queue", "pop max",  n, repeats, pq_setup,
    function(st) pop_max(st$pq)$remaining)

  rows <- bench_one(rows, "base R", "insert",   n, repeats, base_setup,
    function(st) list(values = c(st$v, "val_new"), priorities = c(st$p, 0.5)))
  rows <- bench_one(rows, "base R", "peek min", n, repeats, base_setup,
    function(st) st$v[which.min(st$p)])
  rows <- bench_one(rows, "base R", "pop min",  n, repeats, base_setup,
    function(st) { i <- which.min(st$p); list(values = st$v[-i], priorities = st$p[-i]) })
  rows <- bench_one(rows, "base R", "peek max", n, repeats, base_setup,
    function(st) st$v[which.max(st$p)])
  rows <- bench_one(rows, "base R", "pop max",  n, repeats, base_setup,
    function(st) { i <- which.max(st$p); list(values = st$v[-i], priorities = st$p[-i]) })
}

results_list$pq <- do.call(rbind, as.list(rows))
```
```{r pq-plot, fig.width=9, fig.height=7}
if(!is.null(results_list$pq)) {
  pq_results <- results_list$pq
  pq_results$time_ms <- pq_results$time_us / 1000
  pq_medians <- aggregate(time_us ~ impl + op + n, data = pq_results, FUN = median)
  pq_medians$time_ms <- pq_medians$time_us / 1000

  p_pq <- ggplot(pq_results, aes(x = n, y = time_ms, color = impl)) +
    geom_point(alpha = 0.25, size = 1.2, position = position_jitter(width = 0.03)) +
    geom_line(data = pq_medians, linewidth = 0.6) +
    geom_point(data = pq_medians, size = 1.8) +
    facet_wrap(~ op, scales = "free_y") +
    scale_x_log10(labels = scales::label_comma()) +
    scale_y_log10(labels = scales::label_comma()) +
    labs(
      title = "Priority Queue Operations",
      x = "Number of elements",
      y = "Time (ms)",
      color = "Implementation"
    ) +
    theme_bw() +
    theme(plot.title = element_text(hjust = 0.5), legend.position = "bottom")
  print(p_pq)
} else {
  knitr::asis_output("*Benchmark results not yet generated.*")
}
```




## Interval queries

Insert and three query types (single-point lookup, all-point matches, and
range overlap) on a collection of intervals with integer endpoints. The base R
baseline stores intervals in a `data.frame` and filters with vectorized
comparisons, which is simple and fast for small *n* but O(n) per query.
`interval_index` uses an augmented finger tree for O(log n + k) queries, where
*k* is the number of matches.

```{r ivx-bench, eval=run_benchmarks, cache=FALSE}
ivx_sizes <- c(100, 500, 1000, 5000, 10000, 50000)
rows <- flexseq()

set.seed(123)
max_ivx <- max(ivx_sizes)
all_starts <- sort(sample.int(max_ivx * 10L, max_ivx))
all_widths <- sample.int(100L, max_ivx, replace = TRUE)
all_ends   <- all_starts + all_widths
all_vals   <- sprintf("interval_%06d", seq_len(max_ivx))

qpt   <- all_starts[as.integer(max_ivx / 2)] + 10L
qlo   <- all_starts[as.integer(max_ivx * 0.4)]
qhi   <- all_starts[as.integer(max_ivx * 0.5)]
ins_s <- all_starts[as.integer(max_ivx / 2)]
ins_e <- ins_s + 50L

has_iranges <- requireNamespace("IRanges", quietly = TRUE) &&
  requireNamespace("S4Vectors", quietly = TRUE)

for(n in ivx_sizes) {
  starts <- all_starts[seq_len(n)]
  ends   <- all_ends[seq_len(n)]
  vals   <- all_vals[seq_len(n)]

  ivx_setup  <- function() list(ix = as_interval_index(as.list(vals), start = starts, end = ends, default_query_bounds = "[]"))
  df_setup   <- function() list(df = data.frame(start = starts, end = ends, value = vals, stringsAsFactors = FALSE))

  rows <- bench_one(rows, "interval_index", "insert", n, repeats, ivx_setup,
    function(st) insert(st$ix, "interval_new", ins_s, ins_e))
  rows <- bench_one(rows, "interval_index", "point query", n, repeats, ivx_setup,
    function(st) peek_point(st$ix, qpt, bounds = "[]"))
  rows <- bench_one(rows, "interval_index", "all point matches", n, repeats, ivx_setup,
    function(st) peek_all_point(st$ix, qpt, bounds = "[]"))
  rows <- bench_one(rows, "interval_index", "overlap query", n, repeats, ivx_setup,
    function(st) peek_all_overlaps(st$ix, qlo, qhi, bounds = "[]"))

  rows <- bench_one(rows, "base R", "insert", n, repeats, df_setup,
    function(st) rbind(st$df, data.frame(start = ins_s, end = ins_e, value = "interval_new", stringsAsFactors = FALSE)))
  rows <- bench_one(rows, "base R", "point query", n, repeats, df_setup,
    function(st) {
      hits <- which(st$df$start <= qpt & qpt <= st$df$end)
      if(length(hits)) st$df$value[hits[1L]] else NULL
    })
  rows <- bench_one(rows, "base R", "all point matches", n, repeats, df_setup,
    function(st) st$df[st$df$start <= qpt & qpt <= st$df$end, , drop = FALSE])
  rows <- bench_one(rows, "base R", "overlap query", n, repeats, df_setup,
    function(st) st$df[st$df$start <= qhi & st$df$end >= qlo, , drop = FALSE])

  if(has_iranges) {
    ir_setup <- function() list(
      ir = IRanges::IRanges(start = starts, end = ends),
      v  = vals
    )

    rows <- bench_one(rows, "IRanges", "insert", n, repeats, ir_setup,
      function(st) list(
        ir = c(st$ir, IRanges::IRanges(start = ins_s, end = ins_e)),
        v  = c(st$v, "interval_new")
      ))
    rows <- bench_one(rows, "IRanges", "point query", n, repeats, ir_setup,
      function(st) {
        hits <- S4Vectors::subjectHits(IRanges::findOverlaps(IRanges::IRanges(start = qpt, width = 1L), st$ir))
        if(length(hits)) st$v[hits[1L]] else NULL
      })
    rows <- bench_one(rows, "IRanges", "all point matches", n, repeats, ir_setup,
      function(st) {
        st$v[S4Vectors::subjectHits(IRanges::findOverlaps(IRanges::IRanges(start = qpt, width = 1L), st$ir))]
      })
    rows <- bench_one(rows, "IRanges", "overlap query", n, repeats, ir_setup,
      function(st) {
        st$v[S4Vectors::subjectHits(IRanges::findOverlaps(IRanges::IRanges(start = qlo, end = qhi), st$ir))]
      })
  }
}

results_list$ivx <- do.call(rbind, as.list(rows))
```
```{r ivx-plot, fig.width=9, fig.height=6}
if(!is.null(results_list$ivx)) {
  ivx_results <- results_list$ivx
  ivx_results$time_ms <- ivx_results$time_us / 1000
  ivx_medians <- aggregate(time_us ~ impl + op + n, data = ivx_results, FUN = median)
  ivx_medians$time_ms <- ivx_medians$time_us / 1000

  p_ivx <- ggplot(ivx_results, aes(x = n, y = time_ms, color = impl)) +
    geom_point(alpha = 0.25, size = 1.2, position = position_jitter(width = 0.03)) +
    geom_line(data = ivx_medians, linewidth = 0.6) +
    geom_point(data = ivx_medians, size = 1.8) +
    facet_wrap(~ op, scales = "free_y") +
    scale_x_log10(labels = scales::label_comma()) +
    scale_y_log10(labels = scales::label_comma()) +
    labs(
      title = "Interval Index Queries",
      x = "Number of elements",
      y = "Time (ms)",
      color = "Implementation"
    ) +
    theme_bw() +
    theme(plot.title = element_text(hjust = 0.5), legend.position = "bottom")
  print(p_ivx)
} else {
  knitr::asis_output("*Benchmark results not yet generated.*")
}
```

```{r save-results, eval=run_benchmarks, include=FALSE}
# Persist this run so subsequent builds (CRAN, pkgdown) render the same
# figures without re-timing. Path resolves from knit_root_dir, which the
# generator script sets to the repo root.
saveRDS(results_list, "inst/extdata/benchmarks.rds")
```

```{r save-figures, eval=save_figures, include=FALSE}
dir.create("paper/figures", showWarnings = FALSE, recursive = TRUE)
if(exists("p_sequence")) ggsave("paper/figures/benchmarks-sequence.pdf", p_sequence, width = 9, height = 7)
if(exists("p_queue"))    ggsave("paper/figures/benchmarks-queue.pdf",    p_queue,    width = 8, height = 4.8)
if(exists("p_pq"))       ggsave("paper/figures/benchmarks-pq.pdf",       p_pq,       width = 9, height = 7)
if(exists("p_ivx"))      ggsave("paper/figures/benchmarks-ivx.pdf",      p_ivx,      width = 9, height = 6)
```
