---
title: "Developer API"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Developer API}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
if(requireNamespace("pkgload", quietly = TRUE)) {
  pkgload::load_all(".", quiet = TRUE)
} else if(requireNamespace("immutables", quietly = TRUE)) {
  library(immutables)
} else {
  stop("Need either installed 'immutables' or the 'pkgload' package to render this vignette.")
}
```

Every immutables structure is built on finger trees annotated with *monoids* —
cached summaries that support O(log n) search and split. This vignette covers
the developer API for defining custom monoids, using predicate-based locate and
split operations, and validating tree invariants. These operations work
uniformly across `flexseq`, `ordered_sequence`, `priority_queue`, and
`interval_index`.


## Measures and monoids

A monoid is defined by three pieces: a binary combining function `f`, an
identity value `i`, and a `measure` function that maps a single leaf entry to
its measure value. The combining function *must* be associative 
(`f(a, b) == f(b, a)`), and the identity must satisfy `f(i, x) == x` and `f(x, i) == x`. 

Measure values typically represent something you would want to aggregate over a 
subset of elements, with different measures and aggregations resulting in unique
properties.

When operating on `flexseq`s,
you can think of a monoid's `f` as being applied left-to-right (a left fold), 
accumulating measure values at each element. The calculation for the first 
element also uses `f`, which is where `i` is used, as the other input. 


```{r eval=FALSE}
# elements:  x1,  x2,  x3
# measures:  m1,  m2,  m3,  ...
# monoid-accumulated measures (infix notation):
#  v1 = i %f% m1
#  v2 = (i %f% m1) %f% m2            = v1 %f% m2
#  v3 = ((i %f% m1) %f% m2) %f% m3   = v2 %f% m3
#  ...
```

This left-to-rightness is only well defined for `flexseq`s (using element order 
defined by the user in construction), `ordered_sequence`s (always ordered by key), 
and `interval_index`s (always ordered by interval start). Priority queue element
order is maintained internally with new elements being added on the right, not
in priority order. There may still be uses for monoids that aggregate priority values. 
This is because we can compute an aggregate 
over any *subsequence*, not just prefixes. Internally elements
are stored as leaf nodes in a tree structure, and aggregating measures upward
can be done purely recursively, and cached for fast access at any level.

This monoid keeps track of the sum of values:

```{r}
sum_monoid <- measure_monoid(
  f = `+`,
  i = 0,
  measure = function(el) el
)
```

Next we attach it to
a structure with `add_monoids()`, which accepts a named list of monoids. 
We can pass `show_custom_monoids = TRUE` to `print()` to see this new information:

```{r}
set.seed(42)
task_times <- runif(20, min = 1, max = 10) |>
  round(digits = 1) |>
  as_flexseq()

x <- add_monoids(task_times, list(sum = sum_monoid))

print(x, show_custom_monoids = TRUE)
```

The `plot_structure()` functions visualizes the internal tree layout. Passing a function to
`node_label` lets us render both leaf values and cached structural measures produced by monoids:

```{r fig.height=4, fig.width=10}
plot(x, node_label = function(node) {
  if(node$type == "Element") sprintf("%.1f\nΣ=%.1f", node$element, node$measures$sum)
  else sprintf("Σ=%.1f", node$measures$sum)
})
```

The tree now caches cumulative sum annotations at every internal node,
enabling $O(log n)$ search by accumulated sum. Setting `overwrite = TRUE`
replaces an existing monoid with the same name.

Each structure type reserves certain monoid names for internal use (e.g. `.size`,
`.named_count`, `.pq_min`). Attempting to add a monoid with a reserved name will
error.


## Leaf entry contract

The provided `measure` function receives a structure-specific *entry* — not just the
stored value. The entry shapes are:

- **`flexseq`**: the stored element directly.
- **`ordered_sequence`**: `list(value, key)`.
- **`priority_queue`**: `list(value, priority)`.
- **`interval_index`**: `list(value, start, end)`.

This means a monoid that works on one structure type may not work on another.
Here is a monoid that sums keys on an `ordered_sequence`:

```{r}
key_sum <- measure_monoid(`+`, 0L, function(entry) as.integer(entry$key))

os <- as_ordered_sequence(c("alice", "bob", "carol"), keys = c(10, 20, 30))
os <- add_monoids(os, list(key_sum = key_sum))

print(os, show_custom_monoids = TRUE)
```


## Locating and splitting by predicate

Predicate-based operations scan accumulated monoid values left-to-right and find
the first point where a predicate transitions from `FALSE` to `TRUE`. The
predicate must be *monotonic*: once it returns `TRUE`, it must stay `TRUE` for
all subsequent accumulated values. This is what enables the $O(log n)$ search for a
single split point.

`locate_by_predicate()` finds the *first* element where the predicate fires,
without splitting the tree. With `include_metadata = TRUE` it returns additional
scan information. The predicate function is applied to accumulated measure
values for a given monoid name; here we find the first element where the accumulated
sum exceeds $25$.

```{r}
loc <- locate_by_predicate(x, function(v) v > 25, "sum", include_metadata = TRUE)
str(loc)
```

The `$found` entry indicates that a node where the predicate fires was found; if
this is false the predicate is `FALSE` for all elements through the end. The `$value`
entry gives the measure of the node where the predicate turned `TRUE`.

The optional metadata includes `left_measure` (accumulated value before the match),
`hit_measure` (the matched element's own accumulated measure), `right_measure` (accumulated
value of everything after), and `index` (1-based position). In this example index
4 is where the first time where the accumulated sum is greater than 25, which we can
verify in the plot above ($9.2 + 9.4 + 3.6 = 22.2$, adding in $8.5$ brings the sum to
$30.7$).

`split_by_predicate()` splits the structure at the predicate point, returning
`$left` and `$right`. The matched element is the first element of `$right`:

```{r}
s <- split_by_predicate(x, function(v) v > 25, "sum")
s$left
s$right
```

`split_around_by_predicate()` is similar but extracts the matched element into a
separate `$value` field:

```{r}
sa <- split_around_by_predicate(x, function(v) v > 25, "sum")
sa$left
sa$value
sa$right
```

The `predicate()` wrapper marks a function as a predicate (optional but documents
intent):

```{r}
p <- predicate(function(v) v >= 30)
locate_by_predicate(x, p, "sum")
```

Both split variants accept an optional `accumulator` argument to start scanning
from a value other than the monoid identity. `split_at()` is a convenience
wrapper that splits by position using the built-in `.size` monoid (which counts nodes;
each leaf node's measure is `1` and it uses the `sum` operator):

```{r}
split_at(x, at = 5)
```


## Example: DNA prefix search

Custom monoids and predicate splits combine naturally for index-style queries.
Here we build a persistent subsequence index over a DNA string and use it to find
all positions where a query pattern occurs as a prefix.

The idea: store each position's subsequence as a key in an `ordered_sequence`,
so entries are sorted lexicographically. Then a prefix search becomes two splits —
one to find where the prefix range begins, and one to find where it ends.

```{r}
sequence <- "ACGCGCTCGCGCTTGACGCGCCAG"
query <- "CGCGC"
```

Define a monoid that tracks the maximum key (subsequence string) seen so far. The
built-in `.oms_max_key` monoid already tracks this, but we define our own to
illustrate the pattern:

```{r}
max_seq <- measure_monoid(max, -Inf, function(entry) entry$key)

subseqs <- ordered_sequence()
subseqs <- add_monoids(subseqs, list(max_seq = max_seq))
```

Index every position in the DNA string by its length-10 subsequence:

```{r}
for(i in 1:nchar(sequence)) {
  subseq <- substr(sequence, i, i + 10)
  subseqs <- insert(subseqs, i, key = subseq)
}

subseqs
```

Now search for all subsequences starting with the query. First, split to find
where keys `>=` the query begin:

```{r}
first_split <- split_by_predicate(subseqs, function(m) query <= m, "max_seq")
first_split$left
first_split$right
```

If the first key in `$right` starts with the query, split again to find where
matching keys end. The second predicate fires when the key is strictly greater
than the query *and* no longer starts with it:

```{r}
if(length(first_split$right) > 0) {
  first_key <- min_key(first_split$right)

  if(startsWith(first_key, query)) {
    second_split <- split_by_predicate(
      first_split$right,
      function(m) query < m & !startsWith(m, query),
      "max_seq"
    )
    matches <- second_split$left
  }
}

matches
```

Extract the original string positions from the matching entries:

```{r}
positions <- fapply(matches, function(value, key) value) |> unlist()
positions

# verify: each position is a prefix match
for(pos in positions) {
  cat(pos, ":", substr(sequence, pos, pos + nchar(query) - 1), "\n")
}
```


## Validation helpers

`validate_tree()` checks structural invariants (monoid consistency, node
structure) across the entire tree. `validate_name_state()` verifies that a tree
is either fully unnamed or fully named with unique, non-empty names. Both are
O(n) and intended for debugging and tests, not hot paths.

```{r}
validate_tree(x)
validate_name_state(flexseq(a = 1, b = 2, c = 3))
```
