---
title: "Developer API"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Developer API}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
if(requireNamespace("immutables", quietly = TRUE)) {
  library(immutables)
} else if(requireNamespace("pkgload", quietly = TRUE)) {
  pkgload::load_all(".", quiet = TRUE)
} else {
  stop("Need either installed 'immutables' or the 'pkgload' package to render this vignette.")
}
```

## Scope

This vignette sketches low-level APIs used to extend/query immutable
structures:

- `measure_monoid()`
- `add_monoids()`
- `predicate()`
- `locate_by_predicate()`
- `split_around_by_predicate()` / `split_by_predicate()`
- `validate_tree()` / `validate_name_state()`

## Define and attach a monoid

```{r}
x <- flexseq("a", "bb", "ccc", "dddd")

width_monoid <- measure_monoid(
  f = `+`,
  i = 0L,
  measure = function(el) nchar(el)
)

x2 <- add_monoids(x, list(width = width_monoid))
attr(x2, "measures")$width
```

## Locate with predicate scans

```{r}
p <- predicate(function(v) v >= 5L)

loc <- locate_by_predicate(x2, p, "width", include_metadata = TRUE)
loc$found
loc$elem
loc$metadata
```

## Split with predicate scans

```{r}
s1 <- split_around_by_predicate(x2, function(v) v >= 5L, "width")
s1$left
s1$elem
s1$right

s2 <- split_by_predicate(x2, function(v) v >= 5L, "width")
s2$left
s2$right
```

## Invariant checks

These are intended for debugging and tests.

```{r}
validate_tree(x2)
validate_name_state(as_flexseq(list(a = 1, b = 2)))
```
