---
title: "flexseq"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{flexseq}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  error    = FALSE   # default, but make it explicit: errors halt the build
)

options(
  warn = 2,                     # warnings become errors
  warnPartialMatchDollar = TRUE, # $elem matching $element → warning → error
  warnPartialMatchArgs   = TRUE,
  warnPartialMatchAttr   = TRUE
)

if(requireNamespace("immutables", quietly = TRUE)) {
  library(immutables)
} else if(requireNamespace("pkgload", quietly = TRUE)) {
  pkgload::load_all(".", quiet = TRUE)
} else {
  stop("Need either installed 'immutables' or the 'pkgload' package to render this vignette.")
}
```


## `flexseq` Basics

`flexseq` is a persistent sequence type, with many of the same features as an R `list()`, including storing arbitrary items.
All updates return a new object and keep the original unchanged.

```{r}
x <- flexseq(1, 2, 3)
x

x2 <- as_flexseq(letters[1:5])
x2
```
## Indexing and Slicing, Names

As the output above suggests, they may be indexed and subsetted like lists. Flexseqs may be named, but if any are named they must all have unique non-`NULL` names. Large named flexseqs are somewhat slower than unnamed ones; for efficient 
key/value stores consider using an `ordered_sequence()`.

```{r}
x <- flexseq(a = 1, b = 2, c = 3)
x
x[c("c", "b")]

x$b <- NULL # delete b
x
```
## End operations

Flexseqs support pushing new elements onto the front or back of the sequences, peeking at the front or back (i.e., getting a copy of the first or last item), and popping from the front or back, which returns the removed element and a copy of the flexseq input with that item removed.

```{r}
x <- as_flexseq(4:6)

x <- x |>
  push_back(100) |>
  push_front(50)

x

xpopped <- pop_front(x)

xpopped$value
xpopped$remaining
```
## Indexing and slicing

In general indexing and slicing works as it does with lists: `[]` returns a sublist indexed by integer (by position), character vector 
(by name if named), or logical vector (by inclusion).

Note that these operations are generally expensive, $O(k\log n)$ where
$k$ is the length of the indexing vector. These operations are faster
in base, see the Benchmarks vignette.

Note that vectorized `<`, `>`, `==`, et al. are not yet implemented.

```{r}
x <- as_flexseq(letters[1:6])

x[[3]]

x[c(1, 3, 5)]
```
## Transforming and combining

`c()` concatenates flexseqs, and `fapply()` applies a function to each element of one returning the result as a flexseq, analygous to `lapply()` for R lists.
```{r}
x <- as_flexseq(4:6)
x2 <- as_flexseq(8:10)

c(x, x2)

x <- as_flexseq(1:3)
fapply(x, function(el) el * 10)
```
