---
title: "Benchmarking flexseq Against Base R and rstackdeque"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Benchmarking flexseq Against Base R and rstackdeque}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
if(requireNamespace("pkgload", quietly = TRUE)) {
  pkgload::load_all(".", quiet = TRUE)
} else if(requireNamespace("immutables", quietly = TRUE)) {
  library(immutables)
} else {
  stop("Need either installed 'immutables' or the 'pkgload' package to render this vignette.")
}
source(file.path("helpers", "flexseq_benchmark_helpers.R"), local = TRUE)
run_benchmarks <- flexseq_benchmark_should_run()
bench_sizes <- flexseq_benchmark_default_sizes()
bench_repeats <- flexseq_benchmark_default_repeats()
```

## Method

This article adapts the operation-by-operation style used in Barry (2018),
while narrowing the comparison set to `flexseq`, base R lists, and
`rstackdeque`.

- Each timing cell builds a fresh object of size `n`, then times one operation.
- Sequence timings compare `append`, middle `replace`, middle `get`, and middle
  `remove`.
- Queue timings compare FIFO `enqueue` at the back and `dequeue` from the front.
- The reduced article grid uses sizes `r paste(bench_sizes, collapse = ", ")`
  with `r bench_repeats` repeats per cell.
- Middle reads use `[[` for both `flexseq` and base lists; middle removal times
  the updated sequence (`pop_at(... )$remaining` for `flexseq`).

```{r benchmark-note, eval = !run_benchmarks}
cat(
  paste(
    "Live benchmark chunks are skipped by default.",
    "Set IMMUTABLES_RUN_BENCHMARK_VIGNETTE=1 before rendering to run the reduced grid and regenerate the plots."
  )
)
```

```{r run-benchmarks, eval = run_benchmarks, cache = FALSE}
bench_results <- run_flexseq_benchmark_vignette(
  sizes = bench_sizes,
  repeats = bench_repeats
)
```

## Sequence operations

```{r sequence-plot, eval = run_benchmarks, fig.width = 8, fig.height = 5.5}
bench_results$sequence_plot
```

## Queue/dequeue operations

```{r queue-plot, eval = run_benchmarks, fig.width = 8, fig.height = 4.8}
bench_results$queue_plot
```

## Notes

- `flexseq` sequence updates stay competitive while keeping persistence
  semantics explicit.
- Queue timings use the same FIFO direction across implementations:
  `push_back()`/`pop_front()` for `flexseq`, `insert_back()`/`without_front()`
  for `rstackdeque`, and append/front-removal for base lists.
- These figures are meant for article exposition rather than internal
  regression tracking.
