---
title: "immutables"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{immutables}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
if(requireNamespace("immutables", quietly = TRUE)) {
  library(immutables)
} else if(requireNamespace("pkgload", quietly = TRUE)) {
  pkgload::load_all(".", quiet = TRUE)
} else {
  stop("Need either installed 'immutables' or the 'pkgload' package to render this vignette.")
}
```

## Overview

`immutables` provides four fast data structures:

`flexseq()` - Provides list-like operations including indexed and named element
   access, push/pop/peek from either end for double-ended queue behavior, 
   insertion, splitting, and concatenation.

`priority_queue()` - Associates items with priority values and provides
  min and max peek/pop by priority and fast insertion.

`ordered_sequence()` - Associates items with key values and keeps the elements in
  sorted order by key. These may be similarly be inserted/popped/peeked by key
  value as well as position. Keys may be duplicated, with first-in-first-out 
  order within key groups.

`interval_index()` - Stores items associated with interval ranges, supporting
  point and interval overlap/containment/subsumption queries. Items are kept
  in interval-start-order.
  
Built on monoid-annotated finger trees described by 
[Hinze and Paterson (2006)](https://www.cs.ox.ac.uk/ralf.hinze/publications/FingerTrees.pdf),
they are immutable: operations return modified copies, and are thus 'side-effect free'. 
Their fascinating internal structure ensures most operations are fast, either
constant amortized time or $O(\log(n))$, memory proportional to $n \log(n)$, 
and `immutables` implements core operations in `C++` via `Rcpp` for 
numeric/character/logical key and priority types, while enabling R fallback 
for more complex types such as dates.



## Quick Tour

### flexseq

```{r}
s <- flexseq("a", "b", "c")
s2 <- push_back(s, "d")
```

### priority_queue

```{r}
q <- priority_queue("low", "high", priorities = c(10, 1))
peek_min(q)
pop_min(q)$remaining
```

### ordered_sequence

```{r}
os <- ordered_sequence("one", "two", "two-b", keys = c(1, 2, 2))
peek_key(os, 2)
as.list(peek_all_key(os, 2))
```

### interval_index

```{r}
ix <- interval_index("A", "B", start = c(1, 3), end = c(4, 5))
peek_point(ix, 3)
as.list(peek_all_overlaps(ix, 2, 3))
```

## Choosing a Structure

- Use `flexseq` for general sequence updates and slicing.
- Use `priority_queue` when min/max priority operations are primary.
- Use `ordered_sequence` when key lookup/range queries are primary.
- Use `interval_index` for point/overlap/containment interval relations.

All structures support names using standard R accessors (`seq[c("name1", "name2")]`, `seq[["name"]]`, `seq$name`), but access by name is slower than other operations in some cases. In the case of
ordered sequences and interval indices, return sequences are kept in order, and a warning is displayed if this is different than the requested order. It is generally possible to cast down with `as_flexseq` with

If names are used,
all elements must be non-NULL and unique. Every write-like operation returns a new value and leaves the original unchanged, though assignment with `[<-`, `[[<-`, and `$<-` is possible in  


## Next Articles

- [flexseq](flexseq.html)
- [flexseq Benchmarks](flexseq-benchmarks.html)
- [Priority Queues](priority-queues.html)
- [Ordered Sequences](ordered-sequences.html)
- [Interval Indices](interval-indices.html)
- [Algorithm Demos](algorithm-demos.html)
- [Developer API](developer-api.html)
