---
title: "immutables"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{immutables}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
if(requireNamespace("pkgload", quietly = TRUE)) {
  pkgload::load_all(".", quiet = TRUE)
} else if(requireNamespace("Immutables", quietly = TRUE)) {
  library(Immutables)
} else {
  stop("Need either installed 'Immutables' or the 'pkgload' package to render this vignette.")
}
```

## Overview

`immutables` provides four fast data structures:

`flexseq()` - Provides list-like operations including indexed and named element
   access, push/pop/peek from either end for double-ended queue behavior, 
   insertion, splitting, and concatenation.

`priority_queue()` - Associates items with priority values and provides
  min and max peek/pop by priority and fast insertion.

`ordered_sequence()` - Associates items with key values and keeps the elements in
  sorted order by key. These may be similarly be inserted/popped/peeked by key
  value as well as position. Keys may be duplicated, with first-in-first-out 
  order within key groups.

`interval_index()` - Stores items associated with interval ranges, supporting
  point and interval overlap/containment/subsumption queries. Items are kept
  in interval-start-order.


## Quick Tour

### flexseq

```{r}
s <- flexseq("a", "b", "c", "d")
s |> push_front("z") |> push_back("e")
s[[3]]
peek_back(s)
pop_front(s)$remaining
insert_at(s, 3, c("x", "y"))
fapply(s, toupper)
c(s, as_flexseq(letters[10:12]))
loop(for (el in s) print(el))
```

### priority_queue

```{r}
q <- priority_queue("task_a", "task_b", "task_c", priorities = c(3, 1, 2)) |>
  insert("task_d", priority = 1, name = "d")
peek_min(q)
pop_max(q)$value
peek_all_min(q)$elements
min_priority(q); max_priority(q)
merge(q, priority_queue("task_e", priorities = 0))
```

### ordered_sequence

```{r}
xs <- ordered_sequence("a1", "b1", "b2", "c1", keys = c(1, 2, 2, 3))
peek_key(xs, 2)
peek_all_key(xs, 2)
count_between(xs, 1, 2)
elements_between(xs, 2, 3, include_to = FALSE)
lower_bound(xs, 2)$index
min_key(xs); max_key(xs)
merge(xs, ordered_sequence("d1", keys = 4))
```

### interval_index

```{r}
ix <- interval_index("A", "B", "C", start = c(1, 2, 4), end = c(3, 4, 5))
peek_point(ix, 2)
peek_all_overlaps(ix, start = 2, end = 5)
peek_all_containing(ix, start = 2, end = 3)
peek_all_point(ix, point = 3, match_at = "end")
min_endpoint(ix); max_endpoint(ix)
merge(ix, interval_index("D", start = 6, end = 8))
```

