---
title: "immutables"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{immutables}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
if(requireNamespace("immutables", quietly = TRUE)) {
  library(immutables)
} else if(requireNamespace("pkgload", quietly = TRUE)) {
  pkgload::load_all(".", quiet = TRUE)
} else {
  stop("Need either installed 'immutables' or the 'pkgload' package to render this vignette.")
}
```

## Overview

`immutables` provides persistent (immutable) data structures:

- `flexseq`: general sequence
- `priority_queue`: queue-oriented priority access
- `ordered_sequence`: key-ordered sequence
- `interval_index`: interval query structure

Every write-like operation returns a new value and leaves the original unchanged.

## Quick Tour

### flexseq

```{r}
s <- flexseq("a", "b", "c")
s2 <- push_back(s, "d")
s
s2
```

### priority_queue

```{r}
q <- priority_queue("low", "high", priorities = c(10, 1))
peek_min(q)
pop_min(q)$remaining
```

### ordered_sequence

```{r}
os <- ordered_sequence("one", "two", "two-b", keys = c(1, 2, 2))
peek_key(os, 2)
as.list(peek_all_key(os, 2))
```

### interval_index

```{r}
ix <- interval_index("A", "B", start = c(1, 3), end = c(4, 5))
peek_point(ix, 3)
as.list(peek_all_overlaps(ix, 2, 3))
```

## Choosing a Structure

- Use `flexseq` for general sequence updates and slicing.
- Use `priority_queue` when min/max priority operations are primary.
- Use `ordered_sequence` when key lookup/range queries are primary.
- Use `interval_index` for point/overlap/containment interval relations.

## Next Articles

- [flexseq](flexseq.html)
- [Priority Queues](priority-queues.html)
- [Ordered Sequences](ordered-sequences.html)
- [Interval Indices](interval-indices.html)
- [Developer API](developer-api.html)
