---
title: "Interval Indices"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Interval Indices}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
if(requireNamespace("pkgload", quietly = TRUE)) {
  pkgload::load_all(".", quiet = TRUE)
} else if(requireNamespace("immutables", quietly = TRUE)) {
  library(immutables)
} else {
  stop("Need either installed 'immutables' or the 'pkgload' package to render this vignette.")
}
```

## Interval index basics

`interval_index` is a persistent structure that associates elements with interval endpoints and supports fast interval-relation queries. Intervals are stored sorted by start endpoint, with stable first-in-first-out (FIFO) ordering for ties.

All operations are persistent, and return modified copies. Peeking returns the stored element (which may be any type), popping returns a list with `$value` containing the stored element, `$start` and `$end` the interval endpoints, and `$remaining` the rest of the index without that element.

```{r}
ix <- interval_index(
  "A", "B", "C",
  start = c(1, 2, 4),
  end = c(3, 4, 5)
)
ix
```

The `as_interval_index()` variant builds an index from a vector or list of elements paired with start/end vectors, useful when endpoints are already in separate vectors.

```{r}
ix2 <- as_interval_index(c("phase1", "phase2", "phase3"), start = c(1, 3, 2), end = c(4, 5, 6))
ix2
```

`as_flexseq(ix)` returns a payload-only `flexseq`; interval bounds and any custom monoids are dropped (see `as.list` to convert and preserve interval metadata).


## Point and interval queries

`peek_point()` returns the *first* (in FIFO order) element whose interval contains a query point; `peek_all_point()` returns all matches as an `interval_index` slice.

```{r pointops}
peek_point(ix, point = 2)
peek_all_point(ix, point = 2)
```

`pop_point()` removes and returns the first match as `$value`/`$start`/`$end`/`$remaining`. Its "all" counterpart `pop_all_point()` returns `$elements` (an interval_index of removed matches) and `$remaining`.

```{r}
res <- pop_point(ix, 2)
res$value
res$start
res$end
res$remaining
```

Interval-vs-interval queries select overlaps of different kinds. First we'll build a new index to query:

```{r}
ix3 <- interval_index(
  "narrow", "wide", "right", "inner",
  start = c(2, 1, 4, 3),
  end   = c(3, 6, 5, 4)
)
ix3
```

`peek_overlaps()` returns any interval that shares at least one point with the query range. `peek_containing()` returns intervals that fully contain the query range. `peek_within()` returns intervals fully contained by the query range.

```{r}
# overlaps [2, 5]: any shared point
peek_all_overlaps(ix3, start = 2, end = 5)

# containing [3, 4]: index interval must enclose query
peek_all_containing(ix3, start = 3, end = 4)

# within [1, 5]: index interval must fit inside query
peek_all_within(ix3, start = 1, end = 5)
```

All relation queries have symmetric `pop_*` and `pop_all_*` variants following the same return contract as `pop_point()` and `pop_all_point()`.


## Boundary modes

The `bounds` parameter controls endpoint inclusion: `"[)"` (default, left-closed right-open), `"[]"` (closed), `"()"` (open), or `"(]"` (left-open right-closed). It can be set at construction time as the index default, or overridden per query.

```{r}
edge <- interval_index("E", start = 1, end = 3, bounds = "[)")

# right endpoint excluded by default
peek_point(edge, 3)

# override to closed bounds for this query
peek_point(edge, 3, bounds = "[]")
```

Note that bounds affect what counts as a valid interval: under `"[)"`, an interval with `start == end` like [2, 2) is empty and will never match any query. Use `"[]"` if you need point intervals where `start == end`.

```{r}
# [2, 2) contains nothing under half-open bounds
pt_open <- interval_index("X", start = 2, end = 2, bounds = "[)")
peek_point(pt_open, point = 2)

# [2, 2] contains exactly the point 2
pt_closed <- interval_index("X", start = 2, end = 2, bounds = "[]")
peek_point(pt_closed, point = 2)
```


## Insertion and endpoint extrema

`insert()` adds an element preserving interval start order.

```{r}
ix4 <- insert(ix, "D", start = 2, end = 6)
ix4
```

`min_endpoint()` and `max_endpoint()` return the current smallest start and largest end endpoints (not the stored elements).

```{r}
min_endpoint(ix)
max_endpoint(ix)
min_endpoint(interval_index())
```


## Empty indices

Query and endpoint helpers are non-throwing on empty indices, and `length()` allows for empty checking.

```{r}
empty_ix <- interval_index()
length(empty_ix)
peek_point(empty_ix, point = 1)
pop_overlaps(empty_ix, start = 1, end = 5)
```


## Named interval indices

Interval indices can carry names, set either at construction or through `as_interval_index()` on a named list. Names support read-only indexing via `[`, `[[`, and `$`; all replacement forms (`[<-`, `[[<-`, `$<-`) error, because index-based assignment may break the ordering invariant. Named and unnamed elements cannot be mixed within one index.

```{r}
ix_named <- as_interval_index(
  setNames(list("alice", "bob", "carol"), c("a", "b", "c")),
  start = c(1, 3, 2),
  end = c(4, 5, 6)
)
ix_named

ix_named[["b"]]
ix_named[c("a", "c")]
ix_named[1]

try(ix_named$a <- "!!")
```


## Transforming

`fapply()` maps a function over elements while preserving intervals and order. The function receives `(value, start, end)`, or `(value, start, end, name)` if it accepts a fourth argument. Endpoints and names are passed in read-only, and the return value replaces the stored element (similar to `lapply()` for named R lists).

```{r}
fapply(ix, function(value, start, end) paste0(value, "[", start, ",", end, "]"))
```
