---
title: "Ordered Sequences"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Ordered Sequences}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
if(requireNamespace("immutables", quietly = TRUE)) {
  library(immutables)
} else if(requireNamespace("pkgload", quietly = TRUE)) {
  pkgload::load_all(".", quiet = TRUE)
} else {
  stop("Need either installed 'immutables' or the 'pkgload' package to render this vignette.")
}
```

## Ordered sequence basics

`ordered_sequence` is a persistent structure that associates elements with keys, typically numeric or character (string). They store elements sorted by key, with stable first-in-first-out (FIFO) behavior for duplicate keys, and provide fast key-based lookup, range queries over keys, and insertion.

All operations are persistent, and return modified copies. Peeking by key returns the stored element (which may be any type), popping returns a list with `$value` containing the stored element, `$key` the key, and `$remaining` the rest of the sequence without that element.

```{r}
xs <- ordered_sequence("a1", "b1", "b2", "c1", keys = c(1, 2, 2, 3))
xs
```

The `as_ordered_sequence()` variant builds a sequence from a vector or list of elements paired with a key vector, useful when keys are already in a separate vector.

```{r}
xs2 <- as_ordered_sequence(c(3, 1, 2, 1), keys = letters[1:4])
xs2
```

`as_flexseq(xs)` returns a payload-only `flexseq`; keys and any custom monoids are dropped (see `as.list` to convert and 
preservekey metadata).


## Peek, pop, insert, and range query by key

Insertion is by key, and *keys may be duplicated*. 

```{r}
seq <- as_ordered_sequence(1:3, keys = letters[1:3])
seq2 <- insert(seq, 10, key = "b")
seq2
```

`pop_key()` removes and returns the first match for a key as `$value`/`$key`/`$remaining`. Its "all" counterpart `pop_all_key()` removes the entire tie run for that key, returning `$elements` (an ordered sequence of removed matches) and `$remaining`.

```{r}
one <- pop_key(seq, key = "b")
one$value
one$key
one$remaining

all_two <- pop_all_key(seq, "b")
all_two$elements
all_two$remaining
```

Keys may be counted, and elements accessed; `peek_key()` returns one element, the
first in insertion order. `peek_all_key()` returns an ordered sequence with all
matching keys.

```{r}
count_key(seq, key = "b")
peek_key(seq, key = "b")
peek_all_key(seq, key = "b")
```

Counting and accessing elements over query ranges support inclusive/exclusive on both sides (defaulting to `TRUE`).

```{r}
elements_between(seq, from_key = "b", to_key ="c", include_from = TRUE, include_to = TRUE)
count_between(seq, from_key = "b", to_key = "c", include_from = TRUE, include_to = TRUE)

# exclude "b" keys
elements_between(seq, from_key = "b", to_key ="c", include_from = FALSE, include_to = TRUE)
```


## Key boundaries, and extrema

`lower_bound()` finds the first element with key `>=` a query key; `upper_bound()` finds the first with key strictly `>`. Both return a list with `$found`, `$index`, `$value`, and `$key` (`NULL` fields when no match exists). Together they support successor queries ("find the nearest entry at or above this key") and duplicate counting via index arithmetic.

```{r}
seq <- as_ordered_sequence(1:4, keys = c("b", "d", "d", "f"))

lower_bound(seq, key = "d") |> str()
upper_bound(seq, key = "d") |> str()
```

When the query key falls between or outside existing keys, `lower_bound()` returns the next entry at or above, useful for nearest-match lookups. Both return `found = FALSE` when no keys satisfy the condition.

```{r}
# query key below all keys — both point to the first entry
lower_bound(seq, key = "a") |> str()

# query key above all keys — no match
upper_bound(seq, key = "g") |> str()
```

The difference `upper_bound(x, k)$index - lower_bound(x, k)$index` gives the count of entries with key `k` (this is what `count_key()` does internally).

```{r}
upper_bound(seq, key = "d")$index - lower_bound(seq, key = "d")$index
count_key(seq, key = "d")
```


`min_key()` and `max_key()` return the current minimum and maximum *keys* (not the stored elements).

```{r}
min_key(xs)
max_key(xs)
min_key(ordered_sequence())  # NULL when empty
```

## Empty sequences

Key and range helpers are non-throwing on empty sequences, and `length()` allows for empty checking.

```{r}
empty_os <- ordered_sequence()
length(empty_os)
peek_key(empty_os, 1)
count_between(empty_os, 1, 5)
```

## Named ordered sequences

Ordered sequences can carry names, set either at construction or through `as_ordered_sequence()` on a named list. Names and integer positions both support read-only indexing via `[`, `[[`, and `$`. All replacement forms (`[<-`, `[[<-`, `$<-`) error, because
index-based assignment may break the ordering invariant. Ordered sequences may be cast down with `as_flexseq()` or `as.list()`. Named and unnamed elements cannot be mixed within one sequence.

```{r}
xs_named <- as_ordered_sequence(
  setNames(list("alice", "bob", "carol"), c("a", "b", "c")),
  keys = c(3, 1, 2)
)
xs_named

xs_named[["b"]]
xs_named[c("a", "c")]
xs_named[1]            # positional read also works

try(xs_named$a <- "!!")  # replacement blocked
```

## Transforming

`fapply()` maps a function over elements while preserving keys and order. The function receives `(value, key)`, or `(value, key, name)` if it accepts a third argument. Keys and names are passed in read-only, and the return value replaces the stored element at the associated key (and name if named).

```{r}
xs_t <- ordered_sequence("alice", "bob", "carol", keys = c(3, 1, 2))
fapply(xs_t, function(value, key) toupper(value))
```
