---
title: "Priority Queues: Fast Min/Max Access by Priority"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Priority Queues: Fast Min/Max Access by Priority}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---
```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
if(requireNamespace("pkgload", quietly = TRUE)) {
  pkgload::load_all(".", quiet = TRUE)
} else if(requireNamespace("Immutables", quietly = TRUE)) {
  library(Immutables)
} else {
  stop("Need either installed 'Immutables' or the 'pkgload' package to render this vignette.")
}
```


## Priority queue basics

`priority_queue` is a persistent structure that associates elements with priority values, typically numeric or character (string). They provide
fast access to elements by min or max priority, and insertion of new 
elements with given priorities. 

All operations are persistent, and return modified copies. Peeking at the min/max returns the stored element (which may be any type), 
popping returns a list with `$value` containing the stored element, `$priority` the priority,
and `$remaining` the rest of the priority queue without those elements.

```{r}
x <- priority_queue("task_a", "task_b", "task_c", priorities = c(3, 1, 2))
x

peek_min(x)
res <- pop_max(x)

res$value
res$priority
res$remaining
```

The `as_priority_queue()` variant builds a queue from a vector or list of elements paired with a priority vector. useful when priorities are already in a separate vector.

```{r}
q <- as_priority_queue(letters[1:4], priorities = c(3, 1, 2, 1))
q
```

## Priority ties

In the case of multiple identical priorities, peek/pop for min/max return a *single element*, in their original insertion order. Their "all" counterparts (`peek_all_min`, `pop_all_max`, etc.) return two priorities: in `$elements` the elements peeked/popped and in `$remaining` the other elements. `peek_all_max` and `pop_all_max` behave symmetrically on the maximum-priority tie run.
```{r}
y <- priority_queue() |>
  insert("Jones", priority = "B") |>
  insert("Smith", priority = "A") |>
  insert("Adams", priority = "A")

peek_min(y)
peek_all_min(y)

res1 <- pop_min(y)
res1$value
res1$priority
res1$remaining

res2 <- pop_all_min(y)

res2$elements
res2$remaining
```
Priority queues can be converted to R lists with `as.list()`. The `drop_meta` parameter defaulting to `FALSE` controls whether priority values are included.

```{r}
as.list(res2$elements) |> str()
as.list(res2$elements, drop_meta = TRUE) |> str()
```

## Empty queues

Peek/pop helpers are non-throwing on empty queues, `length()` allows for empty checking.

```{r}
empty_q <- priority_queue()
length(empty_q)
peek_min(empty_q)
pop_min(empty_q)
```

## Priority values

`min_priority()` and `max_priority()` return the current extrema *priority* (not the stored element), in O(1) via cached monoid state.

```{r}
q <- priority_queue("a", "b", "c", priorities = c(3, 1, 2))
min_priority(q)
max_priority(q)
min_priority(priority_queue())  # NULL when empty
```

## Named priorities

Priority queues can carry names, set either at construction or on `insert()`. Names support read-only indexing via `[`, `[[`, and `$`. This is the *only* indexing form `priority_queue` supports. Positional indexing and all replacement forms (`[<-`, `[[<-`, `$<-`) error; cast to `as_flexseq()` first to mutate.

```{r}
q <- priority_queue(a = "task-a", b = "task-b", priorities = c(2, 1)) |>
  insert("task-c", priority = 3, name = "c")

q[["b"]]
q[c("a", "c")]

try(q[1])         # positional indexing blocked
try(q$a <- "!!")  # replacement blocked
```

## Transforming, iterating, merging

`fapply()` maps a function over queue elements while preserving priorities. The function receives `(value, priority)`, or `(value, priority, name)` if it accepts a third argument — priorities and names are passed in read-only, and the return value replaces the stored element.

```{r}
q <- priority_queue("alice", "bob", "carol", priorities = c(3, 1, 2))
fapply(q, function(value, priority) toupper(value))
```

`loop()` (re-exported from the **coro** package) walks the queue in priority-ascending order, yielding bare values. Traversal is driven by repeated `pop_min()`, so full traversal is $O(n \log n)$ ($O(\log n)$ per step); ties within equal priority follow insertion order.

```{r}
q <- priority_queue("task_a", "task_b", "task_c", priorities = c(3, 1, 2))
loop(for (v in q) print(v))
```

Use `fapply()` if your callback needs the priority alongside the value, or cast with `as_flexseq()` for insertion-order iteration. Plain `for (v in q)` (without `loop()`) yields raw finger-tree internals rather than elements — always wrap with `loop()`.

`merge(x, y)` combines two priority queues into a new one containing every entry from both, in O(log(min(m, n))). The `.pq_min` / `.pq_max` monoids recompute automatically on the merged tree, so extremum queries work immediately on the result.

```{r}
a <- priority_queue("x", "y", priorities = c(5, 1))
b <- priority_queue("z", priorities = 3)
m <- merge(a, b)
length(m)
peek_min(m)
peek_max(m)
```

Both queues must share the same priority type and monoid set; mismatches error (rather than being silently harmonized). Both inputs are left unmodified.

For full sequence-style operations, cast with `as_flexseq()`. This returns
payload items only; priority metadata and any custom monoids are dropped. To
instead obtain a `flexseq` of entry records (`value` + `priority`), compose
with `as.list()`.

```{r}
y <- priority_queue() |>
  insert("Jones", priority = "B") |>
  insert("Smith", priority = "A") |>
  insert("Adams", priority = "A")

as_flexseq(y)
as_flexseq(as.list(y))
```

## Example: best-first feature selection

Priority queues have many applications, but we'll demonstrate applying them to the problem of feature selection in modeling: given a large number of features, only a few of which are informative, identify a small subset with good performance. We begin by constructing a matrix `X` with 25 potential feature columns, and an outcome `y` dependent only on features 3 and 7:

```{r}
set.seed(100)
n <- 2000
p <- 25
X <- matrix(rnorm(n * p), n, p)

# response depends on features 3 and 7 only
y <- 3.5 * X[, 3] - 4 * X[, 7] + rnorm(n)
```

We'll consider linear models using only main effects, and use Akaike's Information Criterion (AIC) as a measure of performance. The number of potential models ($2^{25}$) is too large to try all possibilities. Rather than iterate feature combinations exhaustively, we'll perform a best-first search: given a "current best" set of features, we'll add remaining features individually, evaluating as we go and storing them for potential later expansion. By keeping feature sets in a priority queue prioritized by resultant AIC, we can select the best-performing set to build on at each iteration. We thus begin with a base model using no features, and the corresponding empty feature set and AIC stored in a `priority_queue`.

```{r}
# a model using no features, initially the "best"
model <- lm(y ~ 1)
best_features <- numeric(0)
best_aic <- AIC(model)

# store the initial model prioritized by its AIC
pq <- priority_queue()
pq <- insert(pq, best_features, priority = best_aic)
```

So long as there are feature sets available in the queue to expand on, we will do so, but limit the total number of considered models to approximately 200 to prevent an exhaustive search. At each iteration we pop the best feature set considered so far and compute the remaining available features. We then add each of those to the current best set in turn, resulting in a new model with corresponding AIC to add to the queue. If one of these performs better than the best so far we record it. Along the way we update the number of models considered.

```{r}
models_considered <- 1

while(length(pq) > 0 && models_considered < 200) {
  # pop the current best-performing set of features
  current_best <- pop_min(pq)

  # replace the queue with the remaining (unpopped) portion
  pq <- current_best$remaining

  # extract the current best features ($value) and AIC ($priority)
  current_best_features <- current_best$value
  current_best_aic <- current_best$priority

  # features we can add to the current best for potential improvement
  available_features <- setdiff(1:ncol(X), current_best_features)

  # add each unused feature one at a time
  for(available_feature in available_features) {
    # add it to the set of best-performing features so far
    selected_features <- c(current_best_features, available_feature)

    # build a model with that updated set
    model <- lm(y ~ X[ , selected_features, drop = FALSE])

    # add the features to the queue prioritized by the AIC they deliver
    pq <- insert(pq, selected_features, priority = AIC(model))

    # if the current model is better than the best so far, record it
    if(AIC(model) < best_aic) {
      best_features <- selected_features
      best_aic <- AIC(model)
    }

    models_considered <- models_considered + 1
  }
}

best_features
```

The best model encountered by AIC includes the two informative features, 7 and 3, along with several noise features that each improve AIC slightly. Potential improvements to this example include ordering features to avoid duplicated testing of feature permutations, more sophisticated stopping criteria based on AIC trend, and alternative prioritization values. The search history can also easily be inspected efficiently by growing a `flexseq` with intermediate information.
