---
title: "Priority Queues"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Priority Queues}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
loaded <- tryCatch({
  pkgload::load_all(quiet = TRUE)
  TRUE
}, error = function(e) FALSE)

if(!loaded) {
  library(immutables)
}
```


## Priority queue basics

`priority_queue` is a persistent structure that associates elements with priority values, typically numeric or character (string). They provide
fast access to elements by min or max priority, and insertion of new 
elements with given priorities. 

All operations are persistent, and return modified copies. Peeking at the min/max returns the stored element (which may be any type), 
popping returnes a list with `$element` containing the stored element
and `$remaining` a priority queue without those elements.
```{r}
x <- priority_queue("task_a", "task_b", "task_c", priorities = c(3, 1, 2))
x

peek_min(x) # "task_b"
res <- pop_max(x)

res$element
res$priority
res$remaining
```
## Priority ties

In the case of multiple identical priorities, peek/pop for min/max return a *single element*, in their original insertion order. Their "all" counterparts (`peek_all_min`, `pop_all_max`, etc.) return two priorities: in `$elements` the elements peeked/popped and in `$remaining` the other elements.
```{r}
y <- priority_queue() |>
  insert("Jones", priority = "B") |>
  insert("Smith", priority = "A") |>
  insert("Adams", priority = "A")

peek_min(y)
peek_all_min(y)

res1 <- pop_min(y)
res1$element
res1$priority
res1$remaining

res2 <- pop_all_min(y)

res2$elements
res2$remaining
```
Priority queues can be converted to R lists with `as.list()`. The `drop_meta` parameter defaulting to `FALSE` controls whether priority values are included.

```{r}
as.list(res2$elements )
as.list(res2$elements, drop_meta = TRUE)
```

## Empty queues

Peek/pop helpers are non-throwing on empty queues, `length()` allows for empty checking.

```{r}
empty_q <- priority_queue()
length(empty_q)
peek_min(empty_q)
pop_min(empty_q)
```
## Casting to `flexseq`

For full sequence-style operations, cast with `as_flexseq()`. By default this
casts to payload items only (`drop_meta = TRUE`); use
`as_flexseq(..., drop_meta = FALSE)` to keep entry records (`value` +
`priority`).


```{r}
library(immutables)

y <- priority_queue() |>
  insert("Jones", priority = "B") |>
  insert("Smith", priority = "A") |>
  insert("Adams", priority = "A")

as_flexseq(y)
as_flexseq(y, drop_meta = FALSE)
```
