---
documentclass: jss
author:
  - name: Shawn T. O'Neil
    affiliation: University of North Carolina at Chapel Hill
    address: >
      Department of Genetics
      120 Mason Farm Road
      5000 D, Genetic Medicine Building
      CB#7264
      UNC-Chapel Hill
      Chapel Hill, NC 27599-7264
    email: \email{shawn@tislab.org}
title:
  formatted: "\\pkg{Immutables}: Fast, Functional Data Structures for R"
  plain:     "Immutables: Fast, Functional Data Structures for R"
  short:     "\\pkg{Immutables} for R"
abstract: >
  We present \pkg{Immutables}, an R package providing side-effect-free data structures
  including: catenable, indexable sequences that also serve as double-ended queues;
  priority queues; key-ordered sequences; and interval indices for range overlap and similar queries.
  Based on monoid-annotated finger trees first described by Hinze and Paterson in 2006, almost all
  operations are fast, either $O(\log(n))$ or amortized constant time. Finally, a developer
  API exposes custom monoids and aggregations for extensibility. We describe the user and
  developer APIs, provide benchmarks against base-\proglang{R} and comparable implementations,
  and demonstrate extension points with a DNA subsequence search example.
  
  
keywords:
  formatted: ["\\proglang{R}", "\\pkg{Immutables}", "persistent data structures", "finger trees", "monoids", "functional programming"]
  plain:     [R, Immutables, persistent data structures, finger trees, monoids, functional programming]
preamble: >
  \usepackage{amsmath}
output:
  rticles::jss_article:
    citation_package: natbib
    includes:
      in_header: preamble.tex
    extra_dependencies: ["float"]
bibliography: references.bib
classoption: nojss
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.pos = "!htb",
  out.width = "\\linewidth"
)
options(prompt = "R> ", continue = "+  ", width = 70, useFancyQuotes = FALSE)

loaded <- tryCatch({
  pkgload::load_all(quiet = TRUE)
  TRUE
}, error = function(e) FALSE)
if(!loaded) library(Immutables)

library(ggplot2)

bench_results <- local({
  path <- system.file("extdata", "benchmarks.rds", package = "Immutables")
  if(nzchar(path) && file.exists(path)) readRDS(path) else list()
})
```

# Introduction

The \proglang{R} programming language is a natural first choice for
statistical analyses, but its ever-growing popularity has diversified
its application to domains as broad as data serialization and parsing,
\citep{Eddelbuettel2016-iq,Ooms2014-hc}, user interfaces
\citep{Jia2022-si}, simulation \citep{Sipos2011-qc},
mobile computing \citep{Zhu2022-vq},
and much more \citep{Atchison2018-ct}. Data structures, or
collections, are foundational building blocks for almost all computing,
providing the features necessary for efficient data storage, retrieval, and
manipulation. Ordinal-indexed arrays, matrices, and lists are ubiquitous in
statistical applications; stacks and queues support fast insertion
and removal from ends and are widely used in state space search and
scheduling applications \citep{zhang1999state,Gautam2015-bl}; key-ordered sequences
provide fast insertion, access, and range queries 
\citep{Idreos2020-zz}; priority queues provide for efficient access by minimum or maximum
priority value and often used in graph algorithms 
\citep{Borodin2010-pl}. Amongst more exotic structures,
interval trees store sets of ranges for overlap-like queries and are widely used in computational geometry
\citep{Lee1984-xw}.

Data structure implementations generally fall in one of two classes:
mutable or immutable (also known as persistent, or side-effect-free). Operations on mutable
structures allow the internal structure or contents to change.
Operations on immutable structures, by contrast, return changed copies,
preserving the original for further use. Purely functional languages
like \proglang{Haskell} and \proglang{ML} endorse persistence, and while mutating data
structures are generally easier to implement and faster, the functional
programming community has invested significant effort in designing fast
immutable implementations for a wide variety of applications
\citep{Okasaki2011-lz}. This immutability is familiar to \proglang{R}
developers, who expect pass-by-value semantics and are encouraged to develop 
side-effect-free functions \citep{Matloff2011-fh}. Most \proglang{R} structures
follow this paradigm, with the exception of index-assignment (e.g.
\code{x[2] <- 4.2}). Even in these cases, \proglang{R}'s
assign-by-value and copy-on-modify semantics ensure safety
(\code{x <- y; x[2] <- 4.2} modifies \code{x} while preserving the
original data in \code{y}).

Unfortunately, native \proglang{R} types do not emphasize
performance; many operations require a full scan or copy of the
data, making them prohibitively expensive for use in highly dynamic or
large-data applications \citep{Gillespie2016-qx}. Attempts have been made
to fill these gaps. We previously published an implementation of
efficient, persistent stack and queue structures based on lazy
evaluation techniques, noted as the only use of lazy structures in a
2019 review of the \proglang{R} ecosystem \citep{O-Neil2015-je,Goel2019-mz}.
In 2018 T. Barry surveyed the state of collection packages,
identifying 11 packages providing set, map, stack, queue, and/or
sequence structures \citep{Barry2018-mv}. Of these, only three
implemented side-effect-free structures: \pkg{rstackdeque}
\citep{O-Neil2015-je}, \pkg{sets} \citep{Meyer2009-vu}, and
\pkg{S4Vectors} \citep{pages2020s4vectors}. The latter two are optimized
for bulk operations as compared to base-\proglang{R} types but, like
\proglang{R} lists and vectors, do not support fast incremental
insert/remove operations. Notably, \pkg{S4Vectors} is designed for
extensibility and is widely used in the Bioconductor ecosystem,
including support for interval ranges via \pkg{IRanges}
\citep{Lawrence2013-ly}. Barry noted a need for fast, flexible, and
side-effect-free data structures to support advanced use cases in the
\proglang{R} ecosystem \citep{Barry2018-mv}.

Here we describe \pkg{Immutables}, which implements a remarkable data
structure known as monoid-annotated 2-3 finger trees. First described
by Hinze and Paterson \citep{Hinze2006-jv}, these
provide persistent, amortized constant-time access to ends,
$O(\log n)$ search/insertion/removal, and via monoid annotation can
be purposed as other specialized structures. We implement four such
structures: \code{flexseq}s, serving as indexable,
splittable, and concatenable lists with end insertion/removal for
stack and queue operations; priority queues, associating data elements
with priority values for fast min/max access by priority; ordered sequences, keeping
items in order by provided keys; and interval indices, storing items
with ranges for overlap and similar queries. Core operations are
implemented in \proglang{C++} via \pkg{Rcpp} \citep{Eddelbuettel2011-lq} for
speed, and a developer API is provided to extend functionality via
custom monoid annotation.

# Overview

## Flexible sequences

At the core of \pkg{Immutables} is the \code{flexseq} structure.
Much like \proglang{R} lists, these can store
elements of any or mixed types, be named, and be indexed with the
familiar \code{[}, \code{[[}, and \code{$} operators with integer, logical, and
name vectors, including replacement. Elements can be inserted, 
accessed, or removed from either end via peek/pop/push operations in 
amortized constant time for efficient double-ended queue behavior. Additionally,
\code{flexseq}s allow concatenation, splitting by position, and
insertions and removals by position in fast $O(\log n)$ time.

```r
fs1 <- as_flexseq(letters[1:10])
fs2 <- push_back(fs1, "z")
length(fs1)                    # 10
length(fs2)                    # 11

fs3 <- insert_at(fs2, 8, 4.2)  # Immutables structures can store mixed types
length(fs3)                    # 12

fs4 <- fs1[1:5]                # flexseq with "a", "b", "c", "d", "e"

fs5 <- pop_back(fs2)
fs5$value                      # "z"
fs5$remaining                  # flexseq of length 10 (without "z")
```

Other implemented methods include \code{c()}, \code{as.list()},
\code{unlist()}, \code{fapply()} (analogous to \code{lapply()} for
lists), and \code{print()}, mimicking \code{print.list()} but
displaying only the first and last few elements. Native \proglang{R}
\code{for}-loops do not support iterating over elements as a result
of sequences' internal structure, but \code{loop()} is imported from
the \pkg{coro} package to support iterative element access.

While \code{flexseq}s support names, these are somewhat limited
compared to named \proglang{R} lists: if names are used, every
element must have a unique non-\code{NULL} name. Further, the usage
of names degrades worst-case performance for some operations to a
full $O(n)$ scan (for example, when concatenating), though front/back
access remains $O(1)$ amortized in named contexts. Otherwise, all
operations are $O(\log(n))$ worst case, or $O(k \log(n))$ when
removing or accessing $k$ elements. For performance-sensitive applications
requiring named access, ordered sequences (discussed below) can be
associated with character-based names as keys for fast
access.

## Priority queues

The \code{priority_queue} structure stores arbitrary or mixed types, much like
\code{flexseq}, but each element is associated with a priority value.
Rather than efficient front/back access, priority queues provide
efficient insertion and pop/peek by minimum and maximum
priority value. Priorities may be any orderable type (such as dates),
but full \proglang{C++} optimization is available for primitive
numeric, character, and logical priorities. Because multiple elements
may share the same priority, "all" versus "first" access methods are
available: \code{pop\_all\_min(q)} returns an object with two
priority queues, \code{\$elements} containing all elements with
minimum priority and \code{\$remaining} with the rest. 
Conversely, \code{pop\_min(q)} returns the 
first entry (by insertion order) with minimum priority
in \code{$value}, and the rest of the priority queue in
\code{$remaining}. Symmetric \code{_max()} operations are available with 
the same performance characteristics.

```r
pq1 <- as_priority_queue(letters[1:6], priorities = c(2, 1, 3, 1, 9, 7))
peek_min(pq1)              # "b"
peek_all_min(pq1)          # priority queue with "b" "d"

pq2 <- insert(pq1, "z", priority = 10)

pq3 <- pop_max(pq2)
pq3$value                  # "z"
pq3$remaining              # priority queue with 6 entries
```

Elements of priority queues are not stored in priority order, accordingly
priority queues only support indexing and replacement with \code{[}, \code{[[},
and \code{$} by name, with the same caveats as for named \code{flexseq}s.
Priority queues and other advanced types (below) may be 
cast down with \code{as\_flexseq()} and \code{as.list()}, optionally preserving
priority-value metadata. Like \code{flexseq}, priority queues may be mapped
with \code{fapply()}; in this case the called function is supplied both the 
stored element and associated priority, and returns a priority queue
with identical priorities on transformed values. Finally, priority queues may be iteratively
accessed in priority order with \code{loop()}, and they additionally 
provide a priority-aware `merge()`.

## Ordered sequences

The \code{ordered_sequence} type associates each element with a key value
and keeps elements in key order. Keys may be any orderable
type, with optimizations for primitive types, and duplicate keys are
kept in insertion order. Ordered sequences support efficient insert,
peek, and pop by key, with "all" versions of peek and pop for
duplicate-key use cases. Ordered sequences also support
key counts and range queries.

```r
os1 <- as_ordered_sequence(letters[1:6], keys = c(20, 10, 30, 10, 90, 70))
peek_key(os1, 10)               # "b"
peek_all_key(os1, 10)           # ordered sequence with "b" "d"
count_key(os1, 10)              # 2
elements_between(os1, 15, 85)   # ordered seq with "a" "c" "f"

os2 <- insert(os1, "z", key = 100)

os3 <- pop_key(os2, 100)
os3$value                       # character "z"
os3$remaining                   # ordered sequence with "a" through "f"
```

As with other structures, ordered sequences may be named (the same
name caveats apply) and indexed with \code{[}, \code{[[}, and
\code{$}. Integer indices may be used with \code{[} and \code{[[},
in which case they refer to elements in key-sorted order. When
\code{[} is used to subset an ordered sequence, the return is
another ordered sequence and elements are kept in their original
order. A warning is produced if this is not the requested order
(e.g. \code{o[c(2,1)]}). Replacement indexing (\code{[<-} et al.) is not
supported to preserve the ordering invariant.

Also supported is \code{fapply()}, here calling
functions accepting elements and keys and returning an ordered sequence 
with transformed elements, \code{loop()} (in key order), 
and key-aware \code{merge()}. 

## Interval indices

Like other structures above, the `interval_index()` type stores arbitrary
elements with list-like semantics, but each is associated with a $[start, end)$
range. Interval indices thus support peeking and popping by query point, as well 
as query interval overlap, containment, and subsumption. Each of these also has
all and single return variants; \code{pop_overlaps(ix, start, end)} removes and 
returns the first (in start order) overlapping range in \code{ix}, 
while \code{pop_all_overlaps(x, start, end)} removes and returns all. By
default, all bounds are treated as left-closed (\code{[)}); this default
can be adjusted at initialization time, or per-query by supplying a \code{bounds}
argument, in which case the interpretation is applied to both the query and 
stored intervals. Interval endpoints are particularly important for sweep-line
algorithms, so point-based queries can be limited to matching over the whole
interval or only at one or both endpoints.

```r
ix1 <- as_interval_index(c("a", "b", "c"), 
                         start = c(1, 2, 4), 
                         end = c(3, 4, 5))
peek_point(ix1, 2)  # character "a"

ix2 <- insert(ix1, "d", start = 5, end = 6)

ix3 <- pop_all_overlaps(ix2, 3, 4.5)
ix3$elements        # interval index with "b" and "c"
ix3$remaining       # interval index with "a" and "d"
```

Beyond \code{fapply()}, \code{merge()}, \code{loop()}, and conversion methods, interval indices 
can also report global minimum and maximum range endpoints.


# Developer API

Structures in \pkg{Immutables} are backed by monoid-annotated finger trees as
originally described by Hinze and Paterson \citep{Hinze2006-jv}. They are 
simultaneously persistent, performant, and flexible, supporting a variety
of container types and operations. \pkg{Immutables} provides a developer API
exposing this functionality. Here we describe the API and implement two
additional structures: first, a simple accumulator over task times to introduce
the API components, and second, a subsequence-search structure over DNA strings
to illustrate their flexibility.

In functional programming, a \emph{monoid} pairs an associative binary function 
with an identity element, for example $(+,0)$. In the parlance of
Hinze and Paterson, a \emph{measure} is a function mapping stored data elements
to values operated on by a monoid. For our accumulator example, we begin
with a base-\proglang{R} list of tasks, each with an \code{id} and estimated
number of required hours to complete:

```r
tasks <- lapply(1:10, 
                function(i) {
                  list(id = letters[i], 
                  hours = rpois(n = 1, lambda = 5))
                }) |>
         as_flexseq()
```

Given elements such as these, a measuring function 
\code{function(el) el$hours} can be used to map tasks to their estimated
completion time, and when paired with the monoid $(+, 0)$ produces
total time across any sequence of tasks. The \code{measure_monoid()}
function collects these together:


```r
total_time_mm <- measure_monoid(`+`, 0, function(el) el$hours)
```

This monoid can then be attached to the \code{tasks} sequence with 
\code{add_monoids()}. Structures can be associated with multiple monoids,
so they are provided as a named list:

```r
tasks2 <- add_monoids(tasks, list(total_time = total_time_mm))
```

Internally, \code{flexseq}s are tree structures with user values stored in leaf
nodes. Their operations are supported by a set of default monoids;
each of \code{priority_queue}, \code{ordered_sequence}, and \code{interval_index}
subclass \code{flexseq} and add additional monoids supporting their unique 
operations. The \code{plot_structure()} function shows the internal internal structure, 
where node labels can be specified as a function of nodes. Each node has a \code{$type}
set to \code{"Element"} for nodes storing user data, these also have an
\code{$element} storing the data itself, and all nodes carry a 
\code{$measures} storing monoid information. The following code illustrates
our annotated tasks sequence in Figure \ref{fig:tasks-structure}.

```r
plot_structure(tasks2, node_label = function(node) {
  if(node$type == "Element") {
    sprintf("ID: %s, H: %d\nTotal: %d", 
            node$element$id, 
            node$element$hours, 
            node$measures$total_time)
  } else {
    sprintf("Sum: %d", node$measures$total_time)
  }
})
```

```{r tasks-structure, echo=FALSE, dev="pdf", fig.cap="Internal structure of a monoid-annotated finger tree. User values are stored at leaves, illustrated with \\code{ID} and \\code{H} (for Hours). Leaves and internal nodes accumulate a \\code{total\\_time} (shown as \\code{Total}) measure using the associated monoid.", fig.align="center", fig.width=8, fig.height=3.5, out.width="0.95\\textwidth"}
set.seed(42)
tasks <- lapply(1:11, 
                function(i) {
                  list(id = letters[i], 
                  hours = rpois(n = 1, lambda = 5))
                }) |>
         as_flexseq()

total_time_mm <- measure_monoid(`+`, 0, function(el) el$hours)

tasks2 <- add_monoids(tasks, list(total_time = total_time_mm))

plot_structure(tasks2, node_label = function(node) {
  if(node$type == "Element") {
    sprintf("ID: %s, H: %d\nTotal: %d", 
            node$element$id, 
            node$element$hours, 
            node$measures$total_time)
  } else {
    sprintf("Total: %d", node$measures$total_time)
  }
})
```

Several features of Figure \ref{fig:tasks-structure} are notable. First, data 
elements are stored in leaves, in the order of construction. Custom orders are
possible via monoid-based operations; \code{ordered_sequence()} implements
these and provides a convenient key-ordered variant. Second, each leaf node
also carries its measure, here the \code{$hours} value as defined in 
\code{total_time_mm}. Third, internal nodes have different types: Deep, Single, 
and so on. 
Finally, internal nodes also have a measure value, produced by applying the
monoid function \code{+} to all children. These values are accumulated beginning
with the monoid identity \code{0}, and because \code{+} is associative measures
can be accurately aggregated and cached recursively.
Understanding these internals is not required for developer use, 
but this complex structure ensures tree balance for $O(\log(n))$ operations overall, 
and that first andlast elements are a maximum of two levels deep for $O(1)$ operations on ends.

The final piece of the monoid-annotation puzzle comes in the form of a binary
\emph{predicate} function, which must be \emph{monotone over the sequence
of accumulated measure values on leaves}. To motivate, suppose we wish to identify
which tasks we can fully complete in a standard 40-hour work week; the answer is 
tasks \code{a} through \code{f} totaling 39 hours. The following code splits
\code{tasks2} into two subsequences: a \code{$left} whose tasks sum to less 
than or equal to 40 hours (which may be empty), and a \code{$right} with the 
remainder (which also may be empty).

```r
split_tasks <- split_by_predicate(tasks2, 
                                  predicate = function(m) m > 40, 
                                  monoid_name = "total_time")

split_tasks$left  # flexseq with tasks a through f
split_tasks$right # flexseq with tasks g through k
```

Functionally, \code{split_by_predicate()} accumulates measure values defined
by the \code{total_time} monoid $(+, 0)$; \code{a} is 
associated with \code{0 + 8}, \code{b} with \code{(0 + 8) + 9}, \code{c} with 
\code{((0 + 8) + 9) + 4}, and so on. The input sequence is split the first time
the provided predicate over these resolves to \code{TRUE}. Notice that internal 
node totals are not used in this accumulation; rather, they are used by 
\code{split_by_predicate()} to identify the split point with a 
single efficient root-to-leaf traversal.
Monotonicity of the predicate (switching from \code{FALSE} to 
\code{TRUE} at most once) is thus necessary for correctness.

## Example: Subsequence Search

Here we'll illustrate how these components can be used to build an efficient 
subsequence-search data structure modeled after a suffix array, a common approach
to finding exact matches in large DNA strings. 

```r
sequence <- "ACGCGCTCGCGCATAGTCGCGCCTG"
query    <- "CGCGC" # goal: find indices where query occurs (2, 8, and 18)
```

For simplicity we assume that query strings are at most 8 characters long, and
store the position of length-8 subsequences in an \code{ordered_sequence()} 
keyed on the subsequence: key \code{ACGCGCTC} will be associated with index \code{1},
\code{CGCGCTCG} with index \code{2}, and so on. However, the ordered sequence will
store items lexicographically by key: the first few elements will be \code{ACGCGCTC} at index
\code{1}, \code{AGTCGCGC} at \code{15}, \code{ATAGTCGC} at \code{13}, and so on.
Consequently, query matches will be found as prefixes of keys in a continuous
run. 

In order to find the endpoints of a matching run we annotate the sequence with a monoid
that tracks the \emph{maximum} key value, with associative function \code{max} and
identity value \code{-Inf}. When used with an ordered sequence, the measuring
function is passed a list containing both the stored \code{$value} (here the
index) and a \code{$key}. To begin, we'll attach this measuring monoid to an 
empty ordered sequence, and then fill it with keyed values; the structure will
ensure the custom monoid is applied as elements are added.


```r
# max-key monoid
max_seq <- measure_monoid(max, i = -Inf, measure = function(entry) entry$key)

# add the monoid to an empty ordered sequence
subseqs <- ordered_sequence()
subseqs <- add_monoids(subseqs, list(max_seq = max_seq))

# fill with indices keyed on length-8 substrings
for(index in 1:nchar(sequence)) {
  subseq <- substr(sequence, index, index + 7)
  subseqs <- insert(subseqs, index, key = subseq)
}
```

With the structure built, we can find the range matching the query. First
we split the sequence using a predicate that triggers on the first key that is
\code{>=} the query. If any is found \code{$right} will be non-empty, and the
first entry in it \emph{may} have a prefix that is an exact match for the query,
indicating a hit. If it does, we can use a second predicate to find the first
entry that is greater than the query and is not an exact prefix match, capturing
in \code{$left} all exact-match locations. At the end we use \code{fapply()} to
extract only the matching indices as a vector.

```r
matches <- ordered_sequence()

first_split <- split_by_predicate(subseqs, function(m) query <= m, "max_seq")

if(length(first_split$right) > 0) {
  right_measures = get_measures(first_split$right, "max_seq")
  first_measure = right_measures[[1]]

  if(startsWith(first_measure, query)) {
    second_split <- split_by_predicate(
      first_split$right,
      function(m) query < m & !startsWith(m, query),
      "max_seq"
    )
    matches <- second_split$left
  }
}

positions <- fapply(matches, function(value, key) value) |> unlist()
positions   # reports [1] 8 18 2
```

While this structure will not outperform optimized suffix-array 
bioinformatics tools which can achieve linear $O(n)$ build time \citep{Puglisi2007-rk}, 
the performance characteristics here are strong, with an $O(n\log(n))$ build time 
and $O(k + \log(n))$ query time where $k$ is the number of matches.



# Benchmarks

As discussed above, few comparable side-effect-free structures are implemented
in R for comparison. Barry et al. compared performance across several non-persistent
structures in addition to persistent, na\"\i ve implementations using
base-\proglang{R} lists, which are less performant than vectors but support complex element types. 
Here we benchmark \pkg{Immutables} structures against similar
baselines, and for interval indices also compare against \pkg{IRanges}, a 
popular persistent interval-range data structure. All operations were timed
with \code{bench::mark} with 10 iterations on newly initialized structures, garbage collecting with
\code{gc()} before each. Because operations dispatched through S3 methods 
incur first-call costs (method resolution, byte-compilation) that dominate timing 
at small $n$, we warm each operation with three untimed calls before measurement. 
Full benchmark code and results can be found in supplementary materials and in the
\emph{Benchmarks} package vignette.

Sequence and queue operations benchmarked include enqueue (via \code{push_back()}),
dequeue (\code{pop_front()}), concatenation (\code{c()}), and operations around 
the middle element: replacement (\code{[[<-}), removal (\code{pop_at()}), 
and splitting (\code{split_at()}). Base-\proglang{R} sequences were implemented as lists,
with index-based access, removal via negative indexing (e.g. \code{l[-1]} for dequeue), 
and concatenation and enqueue backed by \code{c()}.
To reveal scaling dynamics we ran all tests on structures of between
$2^{12}$ (4,096) and $2^{18}$ (262,144) elements. Results in Figure \ref{fig:benchmarks-sequence}
reveal the $O(n)$ nature of operations on base R lists. The $O(\log(n))$ sequence
operations outperform base lists at between $2^{13}$ (for enqueue/dequeue)
and $2^{17}$ (for middle-element operations) elements.


```{r benchmarks-sequence, echo=FALSE, out.width="0.95\\textwidth", fig.align="center", fig.cap="Benchmark timings for sequence operations.", fig.pos="htb"}
knitr::include_graphics("figures/benchmarks-sequence.pdf")
```

Ordered sequences keep elements sorted by a key value, where duplicate keys
are allowed. Like \code{flexseq}, access to ends (minimum and maximum key)
is amortized $O(1)$, and other operations are $O\log(n)$. In addition to
basic insert/peek/pop operations, we benchmarked count key, key range query,
and lower bounding, a common building block. Base-\proglang{R} element-storing lists
were paired with a character key vector, and kept in common sorted order by key. 
Logical \code{[]}-indexing over results of \code{match()}. All operations were
based on the median key, a worst-case scenario for our ordered sequences.
Results in Figure \ref{fig:benchmarks-ordered} show that base-\proglang{R} 
\code{match()}-based lookup and logical indexing is highly optimized, achieving
~200$\mu s$ key lookups for $2^{18}$ elements. Nevertheless, the linear / logarithmic
divide between implementations is evident, and ordered sequences outperform
base-\proglang{R} lookup at between $2^{17}$ (131k) and $2^{20}$ (1m) elements,
with the exception of \code{peek_key()}, where ordered sequences outperform at
$2^{23}$ (8.3m) elements (not shown).

```{r benchmarks-ordered, echo=FALSE, out.width="0.95\\textwidth", fig.align="center", fig.cap="Benchmark timings for ordered sequence operations.", fig.pos="htb"}
knitr::include_graphics("figures/benchmarks-ordered.pdf")
```

For priority queues we tested simple insertion, peeking, and popping the
maximum priority element (we did not test symmetric \code{max}-based operations which have
identical performance characteristics). Base-\proglang{R} priority queues were implemented 
on an R list with corresponding numeric priority vector, using \code{which.max()}
for index identification and \code{[]}-based indexing and removal. 
Priorities were randomly generated with \code{runif()} with default parameters. 
Figure \ref{fig:benchmarks-pq} reveals priority queues to be substantially faster
than the na\"\i ve implementation at $2^{17}$ elements and beyond.


```{r benchmarks-pq, echo=FALSE, out.width="0.95\\textwidth", fig.align="center", fig.cap="Benchmark timings for priority queue operations.", fig.pos="htb"}
knitr::include_graphics("figures/benchmarks-pq.pdf")
```

We compared interval indices against both base-\proglang{R}
and Bioconductor's \pkg{IRanges} implementations, which are also persistent. 
The former were implemented as data frames with \code{value}, \code{start}, 
and \code{end} columns, filtered using standard row \code{[]}-indexing and
logical selection. Insertion was handled by \code{rbind()}, \pkg{IRanges} structures
store only integer start and end values, so all tests were based on integer ranges
and \pkg{IRanges} objects were accompanied with a corresponding \code{value}
character vector. Operations were supported by \code{[]}-indexing with
selection delegated to \code{findOverlaps} provided to 
\code{subjectHits}; removal by overlap used \code{setdiff}
against the result of \code{subjectHits} and the index range, followed by
logical indexing on the values vector and \pkg{IRanges} object. 

Intervals were generated with integer endpoints, as required by \pkg{IRanges}: 
$n$ start coordinates were sampled without replacement from $[1, 10n]$ and 
sorted, each paired with a width drawn uniformly from $[1, 100]$, yielding a 
sparse collection of short ranges that overlap their near neighbors. We tested
insertion of a new interval, point-based queries for first and all overlapping
intervals, range-based all-containing queries, and removal of all intervals
overlapping a query range. Query points were taken from the 
median stored start position, and query windows extended from this median start 
to the $median + 50^{th}$ start location, resulting in ~50 overlaps for overlap
operations.

```{r benchmarks-ivx, echo=FALSE, out.width="0.95\\textwidth", fig.align="center", fig.cap="Benchmark timings for interval index operations.", fig.pos="htb"}
knitr::include_graphics("figures/benchmarks-ivx.pdf")
```

Results in Figure \ref{fig:benchmarks-ivx} show \pkg{IRanges} to be faster
than base-\proglang{R} for insertion and removal operations, but slower for 
read-only queries. Both implementations exhibit $O(n)$ growth, and interval
indices outperform both at between $2^{16}$ elements for read operations and 
$2^{17}$ elements for insertions and removals.

# Conclusion

\proglang{R}'s continued growth in popularity across many application domains beyond
statistics is in part due to its multi-paradigm nature. Object orientation is provided by
multiple class systems, including \code{R6} which formalizes first-class support for 
mutable data types. Loops and \code{if/then/else} constructs are familiar to those coming
from more procedural languages. Still, \proglang{R}'s roots as a functional language
dominate: functions are pass-by-value, and the generic-dispatch \code{S3} object system 
is most commonly used, followed by \code{S4} in the Bioconductor community \citep{Gentleman2004-tc};
both are typically used in a value-oriented style, where updates create new 
objects via copy-on-modify semantics rather than in-place mutation. Seasoned \proglang{R}
programmers avoid dynamically growing or shrinking structures, which can easily 
result in $O(n^2)$ behavior.

The relative lack of performant, immutable collection types in \proglang{R} has been
previously noted, and \citet{Barry2018-mv} call specifically for their development.
\pkg{Immutables} addresses this by drawing on existing methods in purely functional data structures, 
providing fast off-the-shelf implementations of several common data structures, 
as well as a developer API for custom structures via monoid-annotation.

# Computational details {-}

\pkg{Immutables} is available on CRAN at
\url{https://CRAN.R-project.org/package=Immutables} and may be installed with 
\code{install.packages("Immutables")}. All source code is available at
\url{https://github.com/oneilsh/Immutables}, including a \code{generate_publication_results.R}
script in the \code{data-raw/replication} directory to reproduce all benchmark results, alongside
a faster \code{replication-quick.R} that reproduces representative results in minutes. User-facing
documentation is available in package help pages and vignettes, with
\pkg{pkgdown}-formatted documentation available at \url{https://oneilsh.github.io/Immutables}.

The results in this paper were obtained using `r R.version.string` on
`r R.version$platform` with \pkg{Immutables}~`r packageVersion("Immutables")`
and the comparison packages \pkg{IRanges}~`r packageVersion("IRanges")` 
and \pkg{S4Vectors}~`r packageVersion("S4Vectors")`.
Tests were performed on an Apple M3 Max running MacOS 26.4.

# References {-}
