---
documentclass: jss
author:
  - name: Shawn T. O'Neil
    affiliation: TISLab
    address: >
      Translational and Integrative Sciences Lab
    email: \email{shawn@tislab.org}
title:
  formatted: "\\pkg{immutables}: Persistent, Monoid-Annotated Finger Trees for R"
  plain:     "immutables: Persistent, Monoid-Annotated Finger Trees for R"
  short:     "\\pkg{immutables} for R"
abstract: >
  We present \pkg{immutables}, an R package providing persistent
  finger-tree-backed collections (catenable sequences, priority queues,
  ordered sequences, and interval indices) with user-extensible monoid
  annotations. The same predicate-based locate and split primitives power
  every built-in operation and remain accessible to users, enabling
  $O(\log n)$ queries over custom aggregates. We describe the API,
  provide micro-benchmarks against base R and existing packages, and
  demonstrate the extension points with a DNA prefix-search example.
keywords:
  formatted: ["\\proglang{R}", "finger trees", "persistent data structures", "monoids"]
  plain:     [R, finger trees, persistent data structures, monoids]
preamble: >
  \usepackage{amsmath}
output:
  rticles::jss_article:
    citation_package: natbib
    includes:
      in_header: preamble.tex
bibliography: references.bib
classoption: nojss
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.pos = "!htb",
  out.width = "\\linewidth"
)
options(prompt = "R> ", continue = "+  ", width = 70, useFancyQuotes = FALSE)

loaded <- tryCatch({
  pkgload::load_all(quiet = TRUE)
  TRUE
}, error = function(e) FALSE)
if(!loaded) library(immutables)

library(ggplot2)

bench_results <- local({
  path <- system.file("extdata", "benchmarks.rds", package = "immutables")
  if(nzchar(path) && file.exists(path)) readRDS(path) else list()
})
```

# Introduction

Purely functional collections in \proglang{R} are rare: the stock
\code{list} and \code{vector} types are semantically immutable but the
ergonomic and asymptotic penalties for treating them so are severe.
\pkg{immutables} closes that gap by providing a small suite of
persistent collection types backed by 2-3 finger trees \citep{hinze2006finger}
with user-extensible monoid annotations. The package unifies sequence,
queue, priority-queue, ordered-sequence, and interval-index operations
under a single tree representation and a single pair of
predicate-driven primitives.

# The package

\pkg{immutables} exposes four collection types:

- \code{flexseq} --- a general persistent sequence supporting push/pop at
  both ends, indexing, split, and concatenation.
- \code{priority\_queue} --- a min- and max-priority queue.
- \code{ordered\_sequence} --- a key-ordered sequence with successor,
  predecessor, and range queries.
- \code{interval\_index} --- an interval tree supporting point and range
  overlap queries.

Internally each is a finger tree with a specific set of monoid
annotations attached; the operations above delegate to two primitives
--- \code{locate\_by\_predicate()} and \code{split\_by\_predicate()} ---
over those annotations. Users are free to attach their own monoids via
\code{add\_monoids()} and invoke the same primitives for arbitrary
$O(\log n)$ queries.

# Benchmarks

Figures in this section are loaded from benchmark results produced by
the replication script \code{data-raw/generate\_publication\_results.R}
shipped with the package. The absolute numbers are platform-dependent;
the shapes of the curves are the pertinent evidence.

```{r seq-plot, echo=FALSE, fig.cap="Sequence operations.", fig.width=9, fig.height=7}
if(!is.null(bench_results$sequence)) {
  d <- bench_results$sequence
  d$time_ms <- d$time_us / 1000
  sorted_sizes <- sort(unique(d$n))
  pow_labels <- lapply(sorted_sizes, function(s) bquote(2^.(log2(s)))) |> as.character()
  d$n_cat <- factor(d$n, levels = sorted_sizes)
  ggplot(d, aes(x = n_cat, y = time_ms, color = impl)) +
    geom_point(position = position_jitter(width = 0.15, height = 0)) +
    facet_wrap(~ op, scales = "free_y") +
    scale_x_discrete(labels = pow_labels) +
    labs(x = "Number of elements", y = "Time (ms)", color = "Implementation") +
    theme_bw() + theme(legend.position = "bottom")
}
```

```{r pq-plot, echo=FALSE, fig.cap="Priority queue operations.", fig.width=9, fig.height=7}
if(!is.null(bench_results$pq)) {
  d <- bench_results$pq
  d$time_ms <- d$time_us / 1000
  m <- aggregate(time_us ~ impl + op + n, data = d, FUN = median)
  m$time_ms <- m$time_us / 1000
  ggplot(d, aes(x = n, y = time_ms, color = impl)) +
    geom_point(alpha = 0.25, size = 1.2, position = position_jitter(width = 0.03)) +
    geom_line(data = m, linewidth = 0.6) +
    geom_point(data = m, size = 1.8) +
    facet_wrap(~ op, scales = "free_y") +
    scale_x_log10(labels = scales::label_comma()) +
    scale_y_log10(labels = scales::label_comma()) +
    labs(x = "Number of elements", y = "Time (ms)", color = "Implementation") +
    theme_bw() + theme(legend.position = "bottom")
}
```

```{r ivx-plot, echo=FALSE, fig.cap="Interval index queries.", fig.width=9, fig.height=6}
if(!is.null(bench_results$ivx)) {
  d <- bench_results$ivx
  d$time_ms <- d$time_us / 1000
  m <- aggregate(time_us ~ impl + op + n, data = d, FUN = median)
  m$time_ms <- m$time_us / 1000
  ggplot(d, aes(x = n, y = time_ms, color = impl)) +
    geom_point(alpha = 0.25, size = 1.2, position = position_jitter(width = 0.03)) +
    geom_line(data = m, linewidth = 0.6) +
    geom_point(data = m, size = 1.8) +
    facet_wrap(~ op, scales = "free_y") +
    scale_x_log10(labels = scales::label_comma()) +
    scale_y_log10(labels = scales::label_comma()) +
    labs(x = "Number of elements", y = "Time (ms)", color = "Implementation") +
    theme_bw() + theme(legend.position = "bottom")
}
```

# Extension: DNA prefix search

Custom monoids and predicate splits combine naturally for index-style
queries. We build a persistent suffix-prefix index over a DNA string and
use it to find all positions where a query pattern occurs as a prefix.

```{r dna-setup}
sequence <- "ACGCGCTCGCGCATAGTCGCGCCTG"
query <- "CGCGC"
```

Each position in the sequence is stored as a value keyed on a length-10
window starting at that position. Because \code{ordered\_sequence}
stores entries sorted by key, all matches to a given prefix lie in a
contiguous run, so extracting them is two splits.

We define a monoid tracking the maximum key (subsequence) seen so far
and attach it to an empty \code{ordered\_sequence}:

```{r dna-monoid}
max_seq <- measure_monoid(max, i = -Inf, measure = function(entry) entry$key)
subseqs <- ordered_sequence()
subseqs <- add_monoids(subseqs, list(max_seq = max_seq))
```

Index every position by its length-10 subsequence:

```{r dna-index}
for(i in seq_len(nchar(sequence))) {
  subseqs <- insert(subseqs, i, key = substr(sequence, i, i + 10))
}
```

Split at the first entry whose key is at least the query:

```{r dna-first-split}
first_split <- split_by_predicate(subseqs, function(m) query <= m, "max_seq")
```

If the right side is non-empty and its first key starts with the query,
split again at the first key that is strictly greater *and* no longer
starts with the query:

```{r dna-second-split}
matches <- ordered_sequence()
if(length(first_split$right) > 0L) {
  first_measure <- get_measures(first_split$right, "max_seq")[[1]]
  if(startsWith(first_measure, query)) {
    second_split <- split_by_predicate(
      first_split$right,
      function(m) query < m & !startsWith(m, query),
      "max_seq"
    )
    matches <- second_split$left
  }
}
positions <- fapply(matches, function(value, key) value) |> unlist()
positions
```

The two splits run in $O(\log n)$ each; recovering the matching
positions from a $k$-element result run is $O(k)$. No auxiliary index
structure is maintained beyond the cached monoid values attached to the
tree nodes.

# Conclusion

\pkg{immutables} provides the finger-tree vocabulary as a first-class
extension surface rather than hiding it behind closed collection APIs.
The cost of user-defined monoid attachments is small; the benefit is a
uniform path to $O(\log n)$ operations over arbitrary subtree summaries.

# Computational details {-}

The results in this paper were obtained using \proglang{R}~4.3.0
with the \pkg{immutables} package at its current development version.
Replication is via the script
\code{data-raw/generate\_publication\_results.R} shipped with the
package source.

# Acknowledgments {-}

# References {-}
